---
title: "plant_data_handling"
author: "Marcela Aragon"
date: "3/1/2024"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: 
     collapsed: false
     smooth_scroll: true
    df_print: paged
    code_folding: hide
    keep_md: true
editor_options: 
  chunk_output_type: console
---

This script will deal with arranging and cleaning the raw plant data so it's ready to use in R 


# General settings 

## Rmarkdown

```{r global settings, include=FALSE}

#Set working directory to project directory 
setwd("./")
getwd() #ok

#Set global options for r chunks
knitr::opts_chunk$set(
  echo = TRUE,#show code
  fig.width = 9,
  fig.height = 6,
  dpi = 300,
  error=TRUE, #keep knitting even if error appears
  message=FALSE, #don't show messages
  warning=FALSE, #don't show warnings
  collapse=FALSE)

```

## Loading packages

```{r loading packages, echo=TRUE}

library(knitr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(tibble)
library(readxl)

```

## Global settings for plots 

```{r settings for plots}

# Setting up white background for plots
theme_set(theme_bw())

# Set colors for soil_inocula
color_pathway_induced <- c("999999", "E69F00", "#4472C4") #control, JA and SA
color_experiment <- c("#CC3399", "#70AD47") #simulated and real herbivory

# Set axis looks
axis_looks <- theme(axis.text.x = element_text(colour = "black", size = 10,
                                               face = "bold", angle=45, hjust=1))+
  theme(axis.text.y = element_text(colour = "black", size = 12, face = "bold"))+
  theme(axis.title=element_text(size=15, face = "bold"))+
  theme(axis.title.y = element_text(margin = margin(t = 0, r = 20, b = 0, l = 0)))

```

# Biomass

```{r biomass}

#conditioning
simu <-read_excel("./plant_data/raw_data/raw_biomass.xlsx", sheet = "phytohormones") 
real <-read_excel("./plant_data/raw_data/raw_biomass.xlsx", sheet = "real_herbivory") 

colnames(simu) #"Harvest_ID", "Stress", "Shoots" 
colnames(real) #"ID", "Treatment", "Shoot_biomass(g)", "Leaf_area"

simu2 <- simu %>% 
         select("Harvest_ID", "Stress", "Shoots")

real2 <- real %>% 
         select("ID", "Treatment", "Shoot_biomass(g)", "Leaf_area")

#change names
simu3 <- simu2 %>% 
         rename(shoot_biomass=Shoots)%>%
         rename(ID=Harvest_ID)%>%
         mutate(Leaf_area=NA)%>%
         mutate(Experiment="simulated")

real3 <- real2 %>% 
         rename(Stress=Treatment)%>%
         rename(shoot_biomass='Shoot_biomass(g)')%>%
         mutate(Experiment="real")
               
#merge
biomass <- rbind(simu3, real3)
summary(biomass)

#make factors
biomass <- as.data.frame(unclass(biomass),stringsAsFactors=TRUE)
str(biomass) #check

#let's make a new column with plant_pathway_induced
biomass2 <- biomass %>% 
            mutate(plant_pathway_induced=case_when(
                              Stress == "Aphid" ~ "SA",
                              Stress == "Chewer" ~ "JA",
                              Stress == "Control" ~ "Control",
                              Stress == "MeJA" ~ "JA",
                              Stress == "SA" ~ "SA"))

conditioning <-biomass2

#clean
rm(simu, simu2, simu3, real, real2, real3, biomass2, biomass_simu, biomass_simu2, biomass_real)

#feedback
#simulated
simu <-read_excel("./plant_data/raw_data/raw_biomass.xlsx", sheet = "feedback_phytohormones") 

colnames(simu) #"ID", "Block", "Shoots" 

simu2 <- simu %>% 
         select(-c("Plot", "Place", "Treatment", "Column1", "Aphid sample leaf", 
                   "SA", "JA", "ABA", "Column2", "leaf discs (gr)"))
simu3 <- simu2 %>% 
         rename(weight_shoot_wo_leaf_discs = "Sh. Biomass( without leaf discs)",
                weight_eppendorf = "Eppendorf weight", 
                weight_leaf_discs_mg = "Leaf discs (mg)",
                weight_total_biomass_gr = "total biomass(gr)")

simu4 <- simu3 %>% 
         mutate(herbivory=Herbivory) %>% 
         mutate(soil_inocula = 
                case_when(
                         Soil == "Control 1(POC)" ~ "control-conditioned",
                         Soil == "Control 2 (Live)" ~ "non-conditioned",
                         Soil == "MeJA Conditioned" ~ "JA-conditioned",  
                         Soil == "SA conditioned" ~ "SA-conditioned")) %>%
        mutate(Experiment="simulated")%>%
                   mutate(across(soil_inocula, as.factor))%>%
                   mutate(across(herbivory, as.factor))%>%
                   mutate(across(Experiment, as.factor))

simu5 <- simu4 %>% 
          select(-c("Soil", "Herbivory", "weight_eppendorf"))%>% 
          relocate(ID, Block, Experiment, soil_inocula, herbivory)

simu6 <- simu5 %>%
         
         

#change names
simu3 <- simu2 %>% 
         rename(shoot_biomass=Shoots)%>%
         rename(ID=Harvest_ID)%>%
         mutate(Leaf_area=NA)%>%
         mutate(Experiment="simulated")

real3 <- real2 %>% 
         rename(Stress=Treatment)%>%
         rename(shoot_biomass='Shoot_biomass(g)')%>%
         mutate(Experiment="real")
               
#merge
biomass <- rbind(simu3, real3)
summary(biomass)

#make factors
biomass <- as.data.frame(unclass(biomass),stringsAsFactors=TRUE)
str(biomass) #check

#let's make a new column with plant_pathway_induced
biomass2 <- biomass %>% 
            mutate(plant_pathway_induced=case_when(
                              Stress == "Aphid" ~ "SA",
                              Stress == "Chewer" ~ "JA",
                              Stress == "Control" ~ "Control",
                              Stress == "MeJA" ~ "JA",
                              Stress == "SA" ~ "SA"))
biomass <- biomass2



#save 
write.csv(biomass,"./plant_data/csv/biomass.csv", row.names = FALSE)

```


# Phytohormones

```{r phytohormones}

#load data
phytohormones <-read_excel("./plant_data/raw_data/raw_phytohormones.xlsx", skip=3) 

#split first column
phytohormones <- phytohormones %>% 
                 separate(name, c('Plate', 'Plate_place', 'ID', 'treatment',
                                  'Herbivory_original','Experiment_original'), sep = '[-]')
str(phytohormones)
colnames(phytohormones)

#removing weird column '...8'
phytohormones <- phytohormones[ -c(13) ]

#making factors 
phytohormones[c(1:6)] <- lapply(phytohormones[c(1:6)], factor)

#check
levels(phytohormones$treatment)
levels(phytohormones$Herbivory_original)
levels(phytohormones$Experiment_original)

#New columns:
#Phase
phytohormones <- phytohormones %>%  
                   mutate(Phase =
                            case_when(
                              Experiment_original == "Family" ~ "conditioning",
                              Experiment_original == "Haris" ~ "conditioning",
                              Experiment_original == "Herbivory" ~ "feedback",
                              Experiment_original == "Hormones" ~ "feedback")) 


phytohormones2 <- phytohormones %>%  
                   mutate(Experiment = 
                         case_when(
                              Experiment_original == "Family" ~ "simulated",
                              Experiment_original == "Haris" ~ "real_insects",
                              Experiment_original == "Herbivory" ~ "real_insects",
                              Experiment_original == "Hormones" ~ "simulated")) 
  

phytohormones3 <- phytohormones2 %>%  
                   mutate(Stress = 
                         case_when(
                           Herbivory_original == "without" ~ "Control",
                           Herbivory_original == "caterpillar" ~ "Chewer",
                           Herbivory_original == "aphid" ~ "Aphid",
                           Herbivory_original == "mock" ~ "Control",
                           Herbivory_original == "MeJA" ~ "JA",
                           Herbivory_original == "SA" ~ "SA"))
                           

phytohormones4 <- phytohormones3 %>%  
                   mutate(soil_inocula = 
                         case_when(
                           Phase == "conditioning" ~ "NA",
                           treatment == "control_live" ~ "non-conditioned",
                           treatment == "aphid" ~ "SA-conditioned",
                           treatment == "caterpillar" ~ "JA-conditioned",
                           treatment == "MeJA" ~ "JA-conditioned",
                           treatment == "SA" ~ "SA-conditioned",
                           treatment == "control_plant" & Experiment == "simulated" ~ "control-conditioned",
                           treatment == "control_plant" & Experiment == "real_insects" ~ "control-conditioned")) 


phytohormones5 <- phytohormones4 %>% 
                  mutate(plant_pathway_induced=case_when(
                              Phase == "feedback"  ~ "NA",
                              Phase == "conditioning" & Stress == "Chewer" ~ "JA",
                              Phase == "conditioning" & Stress == "Control" ~ "Control",
                              Phase == "conditioning" & Stress == "Aphid" ~ "SA",
                              Phase == "conditioning" & Stress == "JA" ~ "JA",
                              Phase == "conditioning" & Stress == "SA" ~ "SA"))

# re-naming
colnames(phytohormones5)
phytohormones6 <- phytohormones5 %>%  
                   rename(sum_JA_Ile = 'sum JA-Ile',
                          cis_OPDA = 'cis-OPDA',
                          OH_JA = 'OH-JA',
                          OH_JA_Ile = 'OH-JA-Ile',
                          COOH_JA_Ile = 'COOH-JA-Ile',
                          weight_mg='weight (mg)',
                          herbivory='Herbivory_original')   



#making new columns factors
phytohormones6[c(16:20)] <- lapply(phytohormones6[c(16:20)], factor)

#deleting non-necessary columns 
phytohormones7 <- phytohormones6 %>% 
                  select(-c(Experiment_original, Plate, Plate_place))

#making new column 'Phytohormones'. Look out later to remove those with NA's (not-dried samples for instance)
phytohormones8 <- phytohormones7 %>% 
                  mutate(Phytohormones = "1")

#re-arrange
phytohormones9 <- phytohormones8 %>% 
                  relocate(ID,Phase,Experiment,Stress,herbivory,soil_inocula,plant_pathway_induced, Phytohormones)
str(phytohormones9)

phytohormones9[8] <- lapply(phytohormones9[8], factor)

phytohormones <- phytohormones9

#quick check
ggpairs(phytohormones[,c("Experiment","Phase","Stress","soil_inocula", "plant_pathway_induced", "SA")],
        mapping = aes(color = Experiment)) #there is something weird but let's leave it like that for now

#save 
saveRDS(phytohormones, file = "./plant_data/phytohormones.rds")
write.csv(phytohormones,"./plant_data/csv/phytohormones.csv", row.names = FALSE)

#clean
rm(phytohormones1, phytohormones2, phytohormones3, phytohormones4, phytohormones5,
   phytohormones6, phytohormones7, phytohormones8, phytohormones9)
dev.off()

```

Now I just need to be wary of the 'Loading data' section for each script as some names of treatments have changed (i.e. MeJA_treated by MeJA-treated)!

# Glucosinolates 

```{r glucosinolates}

glucosinolates <-read_excel("./plant_data/raw_data/raw_glucosinolates.xlsx", skip=2) 
str(glucosinolates)

#removing extra columns
glucosinolates <- glucosinolates[ -c(11:13) ]

#separate 1st column 
glucosinolates2 <- glucosinolates %>% 
                 separate(Name, c('Plate', 'Plate_place', 'ID', 'treatment',
                                  'Herbivory_original','Experiment_original'), sep = '[-]')

glucosinolates3 <- glucosinolates2 %>%
                   mutate(across(where(is.character), factor))

str(glucosinolates3)

#check
levels(glucosinolates3$treatment)
levels(glucosinolates3$Herbivory_original)
levels(glucosinolates3$Experiment_original)


#New columns:
#Phase
glucosinolates4 <- glucosinolates3 %>%  
                   mutate(Phase =
                            case_when(
                              Experiment_original == "Family" ~ "conditioning",
                              Experiment_original == "Haris" ~ "conditioning",
                              Experiment_original == "Herbivory" ~ "feedback",
                              Experiment_original == "Hormones" ~ "feedback")) %>%
                   mutate(across(Phase, as.factor))
                   

glucosinolates5 <- glucosinolates4 %>%  
                   mutate(Experiment = 
                         case_when(
                              Experiment_original == "Family" ~ "simulated",
                              Experiment_original == "Haris" ~ "real_insects",
                              Experiment_original == "Herbivory" ~ "real_insects",
                              Experiment_original == "Hormones" ~ "simulated")) %>%
                   mutate(across(Experiment, as.factor))
  

glucosinolates6 <- glucosinolates5 %>%  
                   mutate(Stress = 
                         case_when(Herbivory_original == "without" ~ "Control",
                           Herbivory_original == "caterpillar" ~ "Chewer",
                           Herbivory_original == "aphid" ~ "Aphid",
                           Herbivory_original == "mock" ~ "Control",
                           Herbivory_original == "MeJA" ~ "JA",
                           Herbivory_original == "SA" ~ "SA")) %>%
                   mutate(across(Stress, as.factor))
  

glucosinolates7 <- glucosinolates6 %>%  
                   mutate(soil_inocula = 
                         case_when(
                           Phase == "conditioning" ~ "NA",
                           treatment == "control_live" ~ "non-conditioned",
                           treatment == "aphid" ~ "SA-conditioned",
                           treatment == "caterpillar" ~ "JA-conditioned",
                           treatment == "MeJA" ~ "JA-conditioned",
                           treatment == "SA" ~ "SA-conditioned",
                           treatment == "control_plant" & Experiment == "simulated" ~ "control-conditioned",
                           treatment == "control_plant" & Experiment == "real_insects" ~ "control-conditioned")) %>%
                   mutate(across(soil_inocula, as.factor))

glucosinolates8 <- glucosinolates7 %>% 
                  mutate(plant_pathway_induced=case_when(
                              Phase == "feedback"  ~ "NA",
                              Phase == "conditioning" & Stress == "Chewer" ~ "JA",
                              Phase == "conditioning" & Stress == "Control" ~ "Control",
                              Phase == "conditioning" & Stress == "Aphid" ~ "SA",
                              Phase == "conditioning" & Stress == "JA" ~ "JA",
                              Phase == "conditioning" & Stress == "SA" ~ "SA"))%>%
                   mutate(across(plant_pathway_induced, as.factor))

#checking
str(glucosinolates8)
summary(glucosinolates8)

#deleting non-necessary columns 
glucosinolates9 <- glucosinolates8 %>% 
                  select(-c(Experiment_original, Plate, Plate_place))

#making new column 'glucosinolates' & removing those with NA's (not-dried samples for instance)
glucosinolates10 <- glucosinolates9 %>% 
                  mutate(Glucosinolates = "1")%>% 
                  mutate(across(Glucosinolates, as.factor))%>% 
                  rename(weight_mg='weight [mg]',
                         herbivory=Herbivory_original)
                  

#change names of columns by adding Glu_ because as they start with a number it creates problems  
glucosinolates11 <- glucosinolates10 %>%
                  rename_with(.cols = 5:12, function(x){paste0("Glu_", x)}) 

glucosinolates11 <- glucosinolates11 %>%
                  rename(Glu_3MSOP = Glu_3MSOP...4)

#re-arrange
glucosinolates12 <- glucosinolates11 %>% 
                  relocate(ID,Phase,Experiment,Stress,herbivory,soil_inocula,plant_pathway_induced, Glucosinolates)

str(glucosinolates12)

glucosinolates <- glucosinolates12

#quick check
ggpairs(glucosinolates[,c("Experiment","Phase","Stress","soil_inocula", "plant_pathway_induced", "Glu_total")],
        mapping = aes(color = Experiment)) #there is something weird but let's leave it like that for now

#save 
saveRDS(glucosinolates, file = "./plant_data/glucosinolates.rds")
write.csv(glucosinolates,"./plant_data/csv/glucosinolates.csv", row.names = FALSE)

#clean
rm(glucosinolates2, glucosinolates3, glucosinolates4, glucosinolates5, glucosinolates6,
   glucosinolates7, glucosinolates8, glucosinolates9, glucosinolates10, glucosinolates11, glucosinolates12)

```


# Caterpillars 

```{r caterpillars}

#simulated
simu <-read_excel("./plant_data/raw_data/raw_caterpillars.xlsx", sheet = "Hormones")
str(simu)

#removing extra columns
simu <- simu[-c(10)]

#separate 1st column 
simu <- simu %>% 
        separate(Treatment, c('soil_original', 'herbivory_original'), sep = '[-]')

#making factors 
simu[c(1:5)] <- lapply(simu[c(1:5)], factor)
str(simu)

#make it numeric
simu2 <- simu %>% 
         mutate(across(where(is.character), as.numeric))
str(simu2)

#change df 
simu3 <- simu2 %>% 
         mutate(Experiment="simulated")%>%
         mutate(Phase="feedback")%>%
         mutate(across(where(is.character), as.factor))

simu4 <- simu3 %>%  
         mutate(soil_inocula = 
                case_when(
                         soil_original == "Control 1 (POC) " ~ "control-conditioned",
                         soil_original == "Control 2 (Live) " ~ "non-conditioned",
                         soil_original == "MeJA Conditioned " ~ "JA-conditioned",  
                         soil_original == "SA conditioned " ~ "SA-conditioned")) %>%
                   mutate(across(soil_inocula, as.factor))

simu5 <- simu4 %>% 
         mutate(herbivory="chewer")%>%
         mutate(across(herbivory, as.factor))

#deleting non-necessary columns & adding number of caterpillars
simu6 <- simu5 %>% 
           select(-c(Place, soil_original, herbivory_original))%>%
           relocate(ID,Phase,Experiment,herbivory,soil_inocula)%>%
           mutate(caterpillar_number=5-rowSums(is.na(.))) #5 minus the total number of NA's per row

#insects
real <-read_excel("./plant_data/raw_data/raw_caterpillars.xlsx", sheet = "Insects")
str(real)

#separate 1st column 
real <- real %>% 
        separate(Treatment, c('soil_original', 'herbivory_original'), sep = '[-]')

#making factors 
real[c(1:8)] <- lapply(real[c(1:8)], factor)
str(real)

#make it numeric
real2 <- real %>% 
         mutate(across(where(is.character), as.numeric))

#change df 
real3 <- real2 %>% 
         mutate(Experiment="real_insects")%>%
         mutate(Phase="feedback")%>%
         mutate(across(where(is.character), as.factor))

real4 <- real3 %>%  
         mutate(soil_inocula = 
                case_when(
                         soil_original == "Control 1 (POC) " ~ "control-conditioned",
                         soil_original == "Control 2 (Live) " ~ "non-conditioned",
                         soil_original == "Chewer Conditioned " ~ "JA-conditioned",  
                         soil_original == "Aphid conditioned " ~ "SA-conditioned")) %>%
                   mutate(across(soil_inocula, as.factor))

real5 <- real4 %>% 
         mutate(herbivory="chewer")%>%
         mutate(across(herbivory, as.factor))

#deleting non-necessary columns & adding number of caterpillars
real6 <- real5 %>% 
           select(-c(Place, soil_original, herbivory_original, Plot, Soil, Herbivory))%>%
           relocate(ID,Phase,Experiment,herbivory,soil_inocula)%>%
           mutate(caterpillar_number=5-rowSums(is.na(.))) #5 minus the total number of NA's per row

colnames(real6)
colnames(simu6)

#Merge both df's 
caterpillars <- rbind(x=simu6,y=real6)
str(caterpillars)

#save 
saveRDS(caterpillars, file = "./plant_data/caterpillars.rds")
write.csv(caterpillars,"./plant_data/csv/caterpillars.csv", row.names = FALSE)

#clean
rm(real, real2, real3, real4, real5, real6, simu, simu2, simu3, simu4, simu5, simu6)

```


# Aphids

```{r aphids}

simu <-read_excel("./plant_data/raw_data/raw_aphids.xlsx", sheet = "Hormones")
str(simu)

#removing extra columns
simu <- simu[-c(9)]

#separate 1st column 
simu <- simu %>% 
             separate(Treatment, c('soil_original', 'herbivory_original'), sep = '[-]')

#making factors 
simu[c(1:5)] <- lapply(simu[c(1:5)], factor)
simu[6] <- lapply(simu[6], factor)
str(simu)

#in a plant 15 adults were found 

#remove last two rows
simu2 <- simu %>%
         filter(!row_number() %in% c(41,42)) %>%
         droplevels()

#changing column names for aphids 
simu3 <- simu2 %>%  
              rename(dpi_8 = precount) %>%
              rename(dpi_14 = total) %>% 
              mutate(across(c(dpi_8, dpi_14), as.numeric))

simu4 <- simu3 %>% 
         select(-c('Place', 'Aphid sample leaf', 'Rest of plant'))%>%
         mutate(Experiment="simulated")%>%
         mutate(Phase="feedback")%>%
         mutate(herbivory="aphid")%>%
         mutate(across(where(is.character), as.factor))

simu5 <- simu4 %>% 
         mutate(soil_inocula = 
                case_when(
                         soil_original == "Control 1 (POC) " ~ "control-conditioned",
                         soil_original == "Control 2 (Live) " ~ "non-conditioned",
                         soil_original == "MeJA Conditioned " ~ "JA-conditioned",  
                         soil_original == "SA conditioned " ~ "SA-conditioned")) %>%
                   mutate(across(soil_inocula, as.factor))

simu6 <- simu5 %>% 
         select(-c('soil_original', 'herbivory_original'))%>%
         relocate(ID, Block, Phase, Experiment, soil_inocula, herbivory, dpi_8, dpi_14)

str(simu6)


#Insect-Experiment 
real <-read_excel("./plant_data/raw_data/raw_aphids.xlsx", sheet = "Insects")
str(real)

#making factors 
real[c(1:3)] <- lapply(real[c(1:3)], factor)

real2 <- real %>% 
         mutate(soil_inocula = 
                case_when(
                         Soil == "Control 1 (POC)" ~ "control-conditioned",
                         Soil == "Control 2 (Live)" ~ "non-conditioned",
                         Soil == "Chewer conditioned" ~ "JA-conditioned",  
                         Soil == "Aphid Conditioned" ~ "SA-conditioned")) %>%
                   mutate(across(soil_inocula, as.factor))

real3 <- real2 %>% 
         mutate(herbivory="aphids")%>% 
         mutate(Phase="feedback") %>% 
         mutate(Experiment="real_insects")%>% 
         mutate(across(where(is.character), as.factor))

real4 <- real3 %>%
             rename(dpi_8 ='1st count') %>%
             rename(dpi_14 = Total)%>%
             select(-c(Soil, Herbivore))%>%
             relocate(ID, Phase, Experiment, soil_inocula, herbivory, dpi_8, dpi_14)

str(real4)

#getting block and place missing
blocks <-read_excel("./plant_data/raw_data/raw_aphids.xlsx", sheet = "Insects_Block")
str(blocks)

blocks2 <- blocks %>% 
            select(c(ID, Block))
blocks2[c(1:2)] <- lapply(blocks2[c(1:2)], factor)

id_select <- real4$ID

blocks3 <- blocks2 %>% 
           subset(ID %in% id_select)

#add it
real5 <- real4 %>% 
         mutate(Block = blocks3$Block)

#relocate
real6 <- real5 %>% 
         relocate(Block, .after=ID)

#Merge both df's 
aphids <- rbind(x=simu6, y=real6)
rownames(aphids) <- aphids$ID

#save 
saveRDS(aphids, file = "./plant_data/aphids.rds")
write.csv(aphids,"./plant_data/csv/aphids.csv", row.names = FALSE)

#clean
rm(real, real2, real3, real4, real5, real6, simu, simu2, simu3, simu4, simu5, simu6,
   blocks, blocks2, blocks3)

```


