---
title: "Beta diversity"
author: "Marcela Aragon"
date: "31/07/2024"
output: html_document
---

Continue with differential abundance analysis, one per experiment as you've been doing. Follow the same code that you already have, don't get creative!!


This script will deal with calculating beta diversity for 16S and ITS 

#Loading packages 

```{r}

#Load libraries

#microbiome data
library(phyloseq)
library(microbiome)
library(metagMisc)
library(metagenomeSeq)#CSS normalization
library(vegan)
library(ape)

#data handling
library(dplyr)
library(tidyr)
library(tibble)
library(readxl)
library(purrr) #melts df of lists into a single one
library(varhandle)

#stats
library(picante) #faith's Phylogenetic Diversity
library(stats) #stats
library(DHARMa) #check residuals
library(lme4) #run glms
library(fitdistrplus) #checks data distribution
library(emmeans)
library(rempsyc) #tables
library(magrittr) #pipe with emmeans
library(effectsize)
library(scales)#rounds p value
library(umap)
library(EcolUtils) #pairwise Adonis

#ploting
library(ggplot2)
library(viridis)
library(GGally)
library(ggpubr)


```

## Settings for plots
```{r}

#Set plots theme
ggplot2::theme_set(theme_bw())

# Set default axis looks
axis_looks <- theme(axis.text.x = element_text(colour = "black", size = 9,
                                               face = "bold", angle=0, hjust=0.5))+ #text in X
  theme(axis.text.y = element_text(colour = "black", size = 9, face = "bold"))+ #text in y
  theme(axis.title=element_text(size=11, face = "bold"))+ #all texts
  theme(axis.title.y = element_text(margin = ggplot2::margin(t = 0, r = 10, b = 0, l = 0)))+
  theme(panel.grid.major = element_blank(), #no grid
        panel.grid.minor = element_blank()) #no grid
  #theme(legend.position="none")+ #no legend



#Set colors
color_stress <- c("#BFBFBF", "#E69F00", "#4472C4")
color_experiment <- c("#CC3399", "#70AD47")

```

#ITS

```{r}

##Loading data
#ITS
load(file = "./Results/amplicon/pre_procesing/ps_ITS_FunGuild.RData")
ps_ITS

#split by experiment 
ITS_exp <- phyloseq_sep_variable(ps_ITS, variable="Experiment", drop_zeroes = T)
#ITS_exp <- ITS_exp[-3] #let's remove EPG for now

#Apply filtering to remove low prevalence and abundant ASVs to calculate and plot beta-diversity
asvs_to_keep <- lapply(ITS_exp, function(ps){
  
  ra_ps <- transform_sample_counts(ps, function(x) x/sum(x))#change it to relative abundance
  ra_f <- filter_taxa(ra_ps, function(x) mean(x) >0.00001, prune = TRUE)#filter out less than 0.001%
  p_f <- phyloseq_filter_prevalence(ra_ps, prev.trh=0.05, abund.trh=NULL)#filter out not present in <10% samples
  
  asvs <- rownames(p_f@otu_table) #get names of ASVs
  print(length(asvs)) #get how many
  return(asvs)
})

#make a copy
ITS_exp_f <- ITS_exp

#filter out phyloseq object
ITS_exp_f$cond_phytohormones <- subset_taxa(ITS_exp$cond_phytohormones,
                                            ASV %in% asvs_to_keep$cond_phytohormones)

ITS_exp_f$cond_real_insects <- subset_taxa(ITS_exp$cond_real_insects,
                                            ASV %in% asvs_to_keep$cond_real_insects)

ITS_exp_f$EPG <- subset_taxa(ITS_exp$EPG,ASV %in% asvs_to_keep$EPG)

#final check
ITS_exp_f #ok

#Add number of reads kept per sample
ITS_exp_f <- lapply(ITS_exp_f, function(ps){
 ps@sam_data$libsize_abunprev_filter <-sample_sums(ps)
 return(ps)
})

#see how many reads were kept
sum(taxa_sums(ITS_exp_f$cond_phytohormones))/sum(taxa_sums(ITS_exp$cond_phytohormones)) 
sum(sample_sums(ITS_exp_f$cond_phytohormones))/sum(sample_sums(ITS_exp$cond_phytohormones))
mean(taxa_sums(ITS_exp_f$cond_phytohormones))
mean(sample_sums(ITS_exp_f$cond_phytohormones))
#83% phyto with an average of 14,999.47 reads per ASV and 99,498 reads per sample

sum(taxa_sums(ITS_exp_f$cond_real_insects))/sum(taxa_sums(ITS_exp$cond_real_insects)) 
mean(taxa_sums(ITS_exp_f$cond_real_insects))
mean(sample_sums(ITS_exp_f$cond_real_insects))
#89% insects with an average of 31,770.55 reads per ASV and 98,199 reads per sample

#check sparsity through the process
summarize_phyloseq(ps_ITS)[[6]] #94%
lapply(ITS_exp, function(ps) summarize_phyloseq(ps)[[6]])#91-92
lapply(ITS_exp_f, function(ps) summarize_phyloseq(ps)[[6]])#68-73%

#### Do the same but with the full ps ####
#remove EPG
#make a copy
ps_ITS_f <- ps_ITS
ps_ITS_f <- subset_samples(ps_ITS_f, Experiment %in% c("cond_real_insects", "cond_phytohormones"))
ps_ITS_f <- prune_taxa(taxa_sums(ps_ITS_f) >0, ps_ITS_f)

#Get ASVs to keep
ra_ps <- transform_sample_counts(ps_ITS_f, function(x) x/sum(x))#change it to relative abundance
ra_f <- filter_taxa(ra_ps, function(x) mean(x) >0.00001, prune = TRUE)#filter out less than 0.001%
p_f <- phyloseq_filter_prevalence(ra_f, prev.trh=0.05, abund.trh=NULL)#filter out not present in <5% samples
asvs <- rownames(p_f@otu_table) #get names of ASVs
length(asvs) #get how many

#filter out phyloseq object
ps_ITS_f <- subset_taxa(ps_ITS_f, ASV %in% asvs)

#see how many reads were kept
sum(taxa_sums(ps_ITS_f))/sum(taxa_sums(ps_ITS)) 
mean(taxa_sums(ps_ITS_f))
mean(sample_sums(ps_ITS_f))
#51% with final 28,576.28 reads per ASV and 101,947.8 reads per sample 

#clean
rm(asvs_to_keep, ra_ps, ra_f, p_f)

```

Now, normalize data with CSS transformation
```{r}

ITS_CSS <- lapply(ITS_exp_f, function(ps){phyloseq_transform_css(ps)})
lapply(ITS_CSS, function(ps){head(otu_table(ps))}) #ok

#same for single ps
ps_ITS_css <- phyloseq_transform_css(ps_ITS_f)
head(otu_table(ps_ITS_css))#ok

```

##Distances
```{r}

#Full
bray_full <- phyloseq::distance(ps_ITS_css, "bray")

#Unifrac
uni_full <- phyloseq::distance(ps_ITS_css, "unifrac")


#calculate distance matrix by experiment 
#Bray-Curtis
bray <- lapply(ITS_CSS, function(ps){ phyloseq::distance(ps, "bray")})

#Unifrac
uni <- lapply(ITS_CSS, function(ps){ phyloseq::distance(ps, "unifrac")})

```

##Ordination Plots

###Full
```{r}

dist <- list(bray=bray_full, uni=uni_full) #make it a list

#First get a short version of sample data df 
df_info_ITS <- sample_data(ps_ITS) %>% 
               data.frame() %>% 
               dplyr::select(Experiment, plant_pathway_induced)%>%
               mutate(sample_name=rownames(.))%>%
               filter(Experiment %in% c("cond_real_insects", "cond_phytohormones"))

#call function
get_UMAP_coordenates <- function(x){
  
  #dist is a list of distances previously calculated
  set.seed(123)
  umap <- umap(as.matrix(x))
  umap_scores <- as.data.frame(umap$layout)
  colnames(umap_scores) <- c("UMAP1", "UMAP2")
  umap_scores$sample_name <- rownames(umap_scores)
  umap_scores <- umap_scores %>% relocate(sample_name)
  return(umap_scores)
    
}

#get coordinates
df_UMAP <- lapply(dist, function(x){ get_UMAP_coordenates(x)})

#Now merge it with df 
df_UMAP <- lapply(df_UMAP, function(df){
  x <- left_join(df, df_info_ITS, by="sample_name")
  x <- x %>% relocate(sample_name, Experiment, plant_pathway_induced)
})


#call function
plot_UMAP <- function(x,title){
  #x is df 
  #title is the name of the title in between ""
  set.seed(123)
  p <- ggplot(x, aes(x=UMAP1,y=UMAP2, color=plant_pathway_induced)) +
  geom_point(aes(color=plant_pathway_induced, shape=Experiment),
             size=4, alpha=0.85)+ # individual sample values
  #coord_fixed()+ # same axis scaling
  scale_color_manual(values=color_stress)+
  #ggtitle(title)+
  axis_looks+
  theme(legend.position="none")
  return(p)
}

#Plot it
#u.Bray <- plot_UMAP(df_UMAP$bray, "Bray")
u.uni <- plot_UMAP(df_UMAP$uni, "Unifrac")

u.uni + ylim(-3,3.5)

#Make it a single figure 
#ggarrange(u.Bray,u.uni, ncol=2, nrow=1,common.legend = TRUE, legend="bottom")

#save plot
mapply(function(x)
  ggsave(
  plot = last_plot(),
  filename =x,
  path = "./Plots/02_beta_diversity",
  scale = 1.5,
  width = 800,
  height = 750,
  units = "px",
  dpi = 300),
x=c("UMAP_Phyto-Insects_ITS.svg","UMAP_Phyto-Insects_ITS.png"))

```


###Phyto
Focus for now in nMDS and UMAP for Phyto
```{r}

phyto <- ITS_CSS$cond_phytohormones
dist <- list(bray=bray$cond_phytohormones,uni=uni$cond_phytohormones)

#NMDS
#make a function to calculate NMDS coordinates
get_NMDS_coordenates <- function(x,y){
  
  #x is a phyloseq
  #y is a list of distances
  
  
  df_nmds <- lapply(y, function(dist){
    set.seed(123)
    ord <- ordinate(x, "NMDS", distance=dist)
    nMDS <- plot_ordination(x, ord, color="plant_pathway_induced")
    df_NMDS <- nMDS$data
    df_NMDS <- df_NMDS %>% dplyr::select(Experiment, plant_pathway_induced, NMDS1, NMDS2)
    df_NMDS$sample_name <- rownames(df_NMDS)
    df_NMDS <- df_NMDS %>% relocate(sample_name)
    return(df_NMDS)
    })

  #ord will be a list 
  return(df_nmds)
}

#Get coordenates 
df_NMDS <- get_NMDS_coordenates(phyto,dist)

#Now, get centroids 
centroids <- lapply(df_NMDS, function(df){
  aggregate(cbind(NMDS1, NMDS2) ~ plant_pathway_induced, data=df, FUN=mean)})


#Finally, get segments [I can't make it into a function]
s.bray <- merge(df_NMDS$bray,centroids$bray, by.x="plant_pathway_induced", by.y="plant_pathway_induced",
        suffixes=c("","_centroid"))
s.uni <- merge(df_NMDS$uni,centroids$uni, by.x="plant_pathway_induced", by.y="plant_pathway_induced",
        suffixes=c("","_centroid"))

segments <- list(bray=s.bray, uni=s.uni)
rm(s.bray, s.uni)#clean

#Now, plot the NMDS[make a function to not repeat code]
plot_NMDS <- function(x,y,z,title){
  #x is df 
  #y is the centroid df
  #z is segment df
  #title is the name of the title in between ""
  
  color_stress <- c("#BFBFBF", "#E69F00", "#4472C4")
  
  p <- ggplot(x, aes(x=NMDS1,y=NMDS2, color=plant_pathway_induced)) +
  geom_segment(data = z,
               mapping = aes(x=NMDS1, y=NMDS2, 
                             xend = NMDS1_centroid, yend = NMDS2_centroid),
               linewidth=0.3) + # spiders
  geom_point(size=0.9)+ # individual sample values
  geom_point(data = y, size = 5, alpha=0.9) + # centroids
  #coord_fixed()+ # same axis scaling
  scale_color_manual(values=color_stress)+
  ggtitle(title)
  return(p)
}

#Plot it
p.Bray <- plot_NMDS(df_NMDS$bray, centroids$bray, segments$bray, "Bray")
p.uni <- plot_NMDS(df_NMDS$uni, centroids$uni, segments$uni, "Unifrac")


#Make it a single figure 
ggarrange(p.Bray,p.uni, ncol=2, nrow=1,common.legend = TRUE,legend="bottom")

#save plot
mapply(function(x)
  ggsave(
  plot = last_plot(),
  filename =x,
  path = "./Plots/02_beta_diversity",
  scale = 1.8,
  width = 1000,
  height = 600,
  units = "px",
  dpi = 300),
x=c("NMDS_phyto_ITS.svg","NMDS_phyto_ITS.png"))

```

Now, do the same but with UMAP 

```{r}

phyto <- ITS_CSS$cond_phytohormones
dist <- list(bray=bray$cond_phytohormones,uni=uni$cond_phytohormones)

#First get a short version of sample data df 
df_info_ITS <- sample_data(ps_ITS) %>% 
               data.frame() %>% 
               dplyr::select(Experiment, plant_pathway_induced)%>%
               mutate(sample_name=rownames(.))%>%
               filter(Experiment %in% c("cond_real_insects", "cond_phytohormones"))

#make a function to get UMAP coordenates  
get_UMAP_coordenates <- function(x){
  
  #dist is a list of distances previously calculated
  set.seed(123)
  umap <- umap(as.matrix(x))
  umap_scores <- as.data.frame(umap$layout)
  colnames(umap_scores) <- c("UMAP1", "UMAP2")
  umap_scores$sample_name <- rownames(umap_scores)
  umap_scores <- umap_scores %>% relocate(sample_name)
  return(umap_scores)
    
}

#get coordinates
df_UMAP <- lapply(dist, function(x){ get_UMAP_coordenates(x)})

#Now merge it with df 
df_UMAP <- lapply(df_UMAP, function(df){
  x <- left_join(df, df_info_ITS, by="sample_name")
  x <- x %>% relocate(sample_name, Experiment, plant_pathway_induced)
})

#calculate centroids 
centroids_UMAP <- lapply(df_UMAP, function(df){
  aggregate(cbind(UMAP1, UMAP2) ~ plant_pathway_induced, data=df, FUN=mean)})

#calculate segments
s.bray <- merge(df_UMAP$bray,centroids_UMAP$bray, by.x="plant_pathway_induced", by.y="plant_pathway_induced",
        suffixes=c("","_centroid"))
s.uni <- merge(df_UMAP$uni,centroids_UMAP$uni, by.x="plant_pathway_induced", by.y="plant_pathway_induced",
        suffixes=c("","_centroid"))

segments_UMAP <- list(bray=s.bray, uni=s.uni)
rm(s.bray, s.uni)#clean

#Now, plot the UMAP[make a function to not repeat code]
plot_UMAP <- function(x,y,z,title){
  #x is df 
  #y is the centroid df
  #z is segment df
  #title is the name of the title in between ""
  
  p <- ggplot(x, aes(x=UMAP1,y=UMAP2, color=plant_pathway_induced)) +
  geom_segment(data = z, # spiders
               mapping = aes(x=UMAP1, y=UMAP2, 
                             xend = UMAP1_centroid, yend = UMAP2_centroid),
               linewidth=0.3) + 
  geom_point(size=1.5)+ # individual sample values
  geom_point(data = y, size = 6, alpha=0.9) + # centroids
  #coord_fixed()+ # same axis scaling
  scale_color_manual(values=color_stress)+
    #ggtitle(title)+
  axis_looks+
  theme(legend.position="none")
  return(p)
}

#Plot it
#u.Bray <- plot_UMAP(df_UMAP$bray, centroids_UMAP$bray, segments_UMAP$bray, "Bray")
u.uni <- plot_UMAP(df_UMAP$uni, centroids_UMAP$uni, segments_UMAP$uni, "Unifrac")

#Make it a single figure 
#ggarrange(u.Bray,u.uni, ncol=2, nrow=1,common.legend = TRUE,legend="bottom")

#save plot
mapply(function(x)
  ggsave(
  plot = last_plot(),
  filename =x,
  path = "./Plots/02_beta_diversity",
  scale = 1.5,
  width = 780,
  height = 750,
  units = "px",
  dpi = 300),
x=c("UMAP_phyto_ITS.svg","UMAP_phyto_ITS.png"))

#Finally, make a single df for phyto
#makes it a single df adding a column named "distance" with the name of the list
umap_phyto_df <- map_df(segments_UMAP, ~as.data.frame(.x), .id="distance") 
nmds_phyto_df <- map_df(segments, ~as.data.frame(.x), .id="distance") 

```

###Insects
Continue with Insects
```{r}

insects <- ITS_CSS$cond_real_insects
dist <- list(bray=bray$cond_real_insects,uni=uni$cond_real_insects)

#NMDS
#Get coordenates 
df_NMDS <- get_NMDS_coordenates(insects,dist)

#Now, get centroids 
centroids <- lapply(df_NMDS, function(df){
  aggregate(cbind(NMDS1, NMDS2) ~ plant_pathway_induced, data=df, FUN=mean)})

#Finally, get segments [I can't make it into a function]
s.bray <- merge(df_NMDS$bray,centroids$bray, by.x="plant_pathway_induced", by.y="plant_pathway_induced",
        suffixes=c("","_centroid"))
s.uni <- merge(df_NMDS$uni,centroids$uni, by.x="plant_pathway_induced", by.y="plant_pathway_induced",
        suffixes=c("","_centroid"))

segments <- list(bray=s.bray,  uni=s.uni)
rm(s.bray, s.uni)#clean

#Now, plot the NMDS
p.Bray <- plot_NMDS(df_NMDS$bray, centroids$bray, segments$bray, "Bray")
p.uni <- plot_NMDS(df_NMDS$uni, centroids$uni, segments$uni, "Unifrac")

#Make it a single figure 
ggarrange(p.Bray,p.uni, ncol=2, nrow=1,common.legend = TRUE,legend="bottom")

#save plot
mapply(function(x)
  ggsave(
  plot = last_plot(),
  filename =x,
  path = "./Plots/02_beta_diversity",
  scale = 1.5,
  width = 780,
  height = 750,
  units = "px",
  dpi = 300),
x=c("NMDS_insects_ITS.svg","NMDS_insects_ITS.png"))

```

Continue with UMAP

```{r}

#get coordinates
df_UMAP <- lapply(dist, function(x){ get_UMAP_coordenates(x)})

#Now merge it with df 
df_UMAP <- lapply(df_UMAP, function(df){
  x <- left_join(df, df_info_ITS, by="sample_name")
  x <- x %>% relocate(sample_name, Experiment, plant_pathway_induced)
})

#calculate centroids 
centroids_UMAP <- lapply(df_UMAP, function(df){
  aggregate(cbind(UMAP1, UMAP2) ~ plant_pathway_induced, data=df, FUN=mean)})

#calculate segments
s.bray <- merge(df_UMAP$bray,centroids_UMAP$bray, by.x="plant_pathway_induced", by.y="plant_pathway_induced",
        suffixes=c("","_centroid"))
s.uni <- merge(df_UMAP$uni,centroids_UMAP$uni, by.x="plant_pathway_induced", by.y="plant_pathway_induced",
        suffixes=c("","_centroid"))

segments_UMAP <- list(bray=s.bray, uni=s.uni)
rm(s.bray,  s.uni)#clean

#Now, plot the UMAP
#u.Bray <- plot_UMAP(df_UMAP$bray, centroids_UMAP$bray, segments_UMAP$bray, "Bray")
u.uni <- plot_UMAP(df_UMAP$uni, centroids_UMAP$uni, segments_UMAP$uni, "Unifrac")

#Make it a single figure 
#ggarrange(u.Bray,u.uni, ncol=2, nrow=1,common.legend = TRUE,legend="bottom")

#save plot
mapply(function(x)
  ggsave(
  plot = last_plot(),
  filename =x,
  path = "./Plots/02_beta_diversity",
  scale = 1.5,
  width = 780,
  height = 750,
  units = "px",
  dpi = 300),
x=c("UMAP_insects_ITS.svg","UMAP_insects_ITS.png"))

#Finally, make a single df for phyto
#makes it a single df adding a column named "distance" with the name of the list
umap_insects_df <- map_df(segments_UMAP, ~as.data.frame(.x), .id="distance") 
nmds_insects_df <- map_df(segments, ~as.data.frame(.x), .id="distance") 

```

Finally, let's place everything in a single df to save it 
```{r}

#makes it a single df adding a column named "distance" with the name of the list
umap_df <- rbind(umap_phyto_df,umap_insects_df )
nmds_df <- rbind(nmds_phyto_df,nmds_insects_df )

#save it as .csv df
write.csv(umap_df, file="./Data_frames/02_beta_diversity/umap_ITS.csv")
write.csv(nmds_df, file="./Data_frames/02_beta_diversity/nmds_ITS.csv")

```

##PERMANOVA
```{r}

#I need this to test both experiments together 
ps_ITS_css
df_info_ITS

#Is there a difference between the fungal communities among types of induction(Experiment) and treatments?
set.seed(123)

perm_full_bray <- adonis2(formula=phyloseq::distance(ps_ITS_css, "bray") 
                          ~ Experiment +plant_pathway_induced,
                     data=as(sample_data(ps_ITS_css), "data.frame"),
                     permutations = 999, by="margin") #yes

perm_full_uni <- adonis2(formula=phyloseq::distance(ps_ITS_css, "unifrac") 
                          ~ Experiment + plant_pathway_induced,
                     data=as(sample_data(ps_ITS_css), "data.frame"),
                     permutations = 999, by="margin") #yes

#which treatments are different from each other? --> not working 13/09
pairwise_bray <- adonis.pair(dist.mat= distance(ps_ITS_css, "bray"),
                 Factor = as.factor(as(sample_data(ps_ITS_css), "data.frame")$plant_pathway_induced),
                 nper=1000,corr.method="fdr")#Ctrl vs JA

pairwise_uni <- adonis.pair(dist.mat= distance(ps_ITS_css, "unifrac"),
                 Factor = as.factor(as(sample_data(ps_ITS_css), "data.frame")$plant_pathway_induced),
                 nper=1000,corr.method="fdr")#Ctrl vs JA

```

Save results
```{r}

#Permanova
perm_full_bray$distance <- "Bray-Curtis"
perm_full_bray <- perm_full_bray %>% rownames_to_column(var="factor") %>% relocate(distance)

perm_full_uni$distance <- "Unifrac"
perm_full_uni <- perm_full_uni %>% rownames_to_column(var="factor") %>% relocate(distance)

permanova_full <- rbind(perm_full_bray, perm_full_uni)

#save it
table <- nice_table(permanova_full ,
              title = c("Table X. PERMANOVA for the full ITS community"))

flextable::save_as_docx(table,
                        path = "./Tables/02_beta_diversity/full_permanova_ITS_BothExperiments.docx.docx")

```


```{r}
#Pairwise comparisons --> not working 
#make it into a df
pairwise_bray$distance <- "Bray-Curtis"
pairwise_bray <- pairwise_bray %>% relocate(distance)

pairwise_uni$distance <- "Unifrac"
pairwise_uni <- pairwise_uni %>% relocate(distance)

pairwise_full <- rbind(pairwise_bray, pairwise_uni) #merge it

#save it
table <- nice_table(pairwise_full ,
              title = c("Table X. Overall pairwise comparisons between treatments for the full ITS community"))
flextable::save_as_docx(table, path = "./Tables/02_beta_diversity/full_pairwise_ITS.docx")

```


```{r}
#I need this to test both experiments independently
ITS_CSS
bray
uni

ITS_CSS <- ITS_CSS[-3] #remove EPG for now 

#Is there a difference between the fungal communities among treatments?
set.seed(123)

#Bray-Curtis
perm_Byexp_bray <- lapply(ITS_CSS, function(ps){
          set.seed(123)
          perm <- adonis2(formula=phyloseq::distance(ps, "bray") 
                          ~ plant_pathway_induced,
                     data=as(sample_data(ps), "data.frame"),
                     permutations = 999, by="margin")
          perm <- perm %>% rownames_to_column("factor")
          return(perm)
          })


perm_Byexp_bray <- map_df(perm_Byexp_bray, ~as.data.frame(.x), .id="Experiment") 
perm_Byexp_bray <- perm_Byexp_bray %>% mutate(distance="Bray-Curtis") %>% relocate(distance)

#Unifrac
perm_Byexp_uni <- lapply(ITS_CSS, function(ps){
          set.seed(123)
          perm <- adonis2(formula=phyloseq::distance(ps, "unifrac") 
                          ~ plant_pathway_induced,
                     data=as(sample_data(ps), "data.frame"),
                     permutations = 999, by="margin")
          perm <- perm %>% rownames_to_column("factor")
          return(perm)
          })

perm_Byexp_uni <- map_df(perm_Byexp_uni, ~as.data.frame(.x), .id="Experiment") 
perm_Byexp_uni <- perm_Byexp_uni %>% mutate(distance="Unifrac") %>% relocate(distance)

#merge
permanova_Byexp <- rbind(perm_Byexp_bray, perm_Byexp_uni)

#save it
table <- nice_table(permanova_Byexp,
              title = c("Table X. PERMANOVA test by Experiment for the ITS community"))

#flextable::save_as_docx(table, path = "./Tables/02_beta_diversity/permanova_ByExperiment_ITS.docx")

#save df just in case 
write.table(permanova_Byexp, "./Data_frames/02_beta_diversity/permanova_byexperiment_ITS.tsv", quote = F, row.names = F, sep = "\t")

```


```{r}
#Now, get pairwise differences but just for phytohormone experiment
#which treatments are different from each other?
pairwise_bray <- adonis.pair(dist.mat= distance(ITS_CSS$cond_phytohormones, "bray"),
                 Factor =as.factor(as(sample_data(ITS_CSS$cond_phytohormones),
                                      "data.frame")$plant_pathway_induced),
                 nper=1000,corr.method="fdr")#Ctrl vs JA

pairwise_uni <- adonis.pair(dist.mat= distance(ITS_CSS$cond_phytohormones, "unifrac"),
                 Factor = as.factor(as(sample_data(ITS_CSS$cond_phytohormones),
                                       "data.frame")$plant_pathway_induced),
                 nper=1000,corr.method="fdr")#Ctrl vs JA

pairwise_bray$distance <- "Bray-Curtis"
pairwise_bray <- pairwise_bray %>% relocate(distance)

pairwise_uni$distance <- "Unifrac"
pairwise_uni <- pairwise_uni %>% relocate(distance)

pairwise_full <- rbind(pairwise_bray, pairwise_uni) #merge it

#save it
table <- nice_table(pairwise_full ,
              title = c("Table X. Overall pairwise comparisons between treatments of the ITS community for the Phytohormone experiment "))
flextable::save_as_docx(table, path = "./Tables/02_beta_diversity/phyto_pairwise_ITS.docx")

```

```{r}
#clean
#remove everything but functions and colors for plots
rm.all.but(keep=c("color_experiment", "color_stress"), keep_functions = TRUE)
```


```{r}
save.image("./R_Environments/02_beta_diversity/ITS.Rdata")
```



Finished for ITS, continue with 16S

#16S

Lesson learned from ITS; focus on Bray-Curtis and Unifrac


```{r}

##Loading data
#16S
load(file = "./amplicon_data/Phyloseq_objects/ps_16S.RData")
ps_16S

#16S split by experiment
load(file = "./amplicon_data/Phyloseq_objects/ps_16S_split.RData")

```

Apply filtering by abundance and prevalence
```{r}
#Add ASV name in taxonomy table
t <- as.data.frame(tax_table(ps_16S))
t$ASV <- rownames(t)
tax_table(ps_16S) <- tax_table(object = as.matrix(t))

#Add it also to the split ones
ps_16S_split <- lapply(ps_16S_split, function(ps){
  t <- as.data.frame(tax_table(ps))
  t$ASV <- rownames(t)
  tax_table(ps) <- tax_table(object = as.matrix(t))
  return(ps)
})

#Apply filtering to remove low prevalence and abundant ASVs to calculate and plot beta-diversity
asvs_to_keep <- lapply(ps_16S_split, function(ps){
  
  ra_ps <- transform_sample_counts(ps, function(x) x/sum(x))#change it to relative abundance
  ra_f <- filter_taxa(ra_ps, function(x) mean(x) > 0.00001, prune = TRUE)#filter out less than 0.001%
  p_f <- phyloseq_filter_prevalence(ra_f, prev.trh=0.1, abund.trh=NULL)#filter out not present in <10% samples
  
  asvs <- rownames(p_f@otu_table) #get names of ASVs
  print(length(asvs)) #get how many
  return(asvs)
})

#make a copy
Bac_exp_f <- ps_16S_split
lapply(ps_16S_split, function(x) mean(taxa_sums(x)))

#filter out phyloseq object
Bac_exp_f$cond_phytohormones <- subset_taxa(ps_16S_split$cond_phytohormones,
                                            ASV %in% asvs_to_keep$cond_phytohormones)

Bac_exp_f$cond_real_insects <- subset_taxa(ps_16S_split$cond_real_insects,
                                            ASV %in% asvs_to_keep$cond_real_insects)

Bac_exp_f$EPG <- subset_taxa(ps_16S_split$cond_real_insects,
                                            ASV %in% asvs_to_keep$EPG)

#final check
Bac_exp_f #ok

#see how many reads were kept
sum(taxa_sums(Bac_exp_f$cond_phytohormones))/sum(taxa_sums(ps_16S_split$cond_phytohormones)) 
mean(taxa_sums(Bac_exp_f$cond_phytohormones))
mean(sample_sums(Bac_exp_f$cond_phytohormones))
#89% phyto with final average of 1,103 reads per ASV and 83,801.93 reads per sample

sum(taxa_sums(Bac_exp_f$cond_real_insects))/sum(taxa_sums(ps_16S_split$cond_real_insects)) 
mean(taxa_sums(Bac_exp_f$cond_real_insects))
mean(sample_sums(Bac_exp_f$cond_real_insects))
#91% insects final average of 1,459 reads per ASV and 60,110.14 reads per sample

sum(taxa_sums(Bac_exp_f$EPG))/sum(taxa_sums(ps_16S_split$EPG)) 
mean(taxa_sums(Bac_exp_f$EPG))
mean(sample_sums(Bac_exp_f$EPG))
#66% insects final average of 1,006 reads per ASV and 60,177 reads per sample


#check sparsity through the process
summarize_phyloseq(ps_16S)[[6]] #97%
lapply(ps_16S_split, function(ps) summarize_phyloseq(ps)[[6]])#92
lapply(Bac_exp_f, function(ps) summarize_phyloseq(ps)[[6]])#63-64 and 74%

#### Do the same but with the full ps ####
#remove EPG
#make a copy
ps_Bac_f <- ps_16S
ps_Bac_f <- subset_samples(ps_Bac_f, Experiment %in% c("cond_real_insects", "cond_phytohormones"))
ps_Bac_f <- prune_taxa(taxa_sums(ps_Bac_f) >0, ps_Bac_f)

#Get ASVs to keep
ra_ps <- transform_sample_counts(ps_Bac_f, function(x) x/sum(x))#change it to relative abundance
ra_f <- filter_taxa(ra_ps, function(x) mean(x) >0.00001, prune = TRUE)#filter out less than 0.0001%
p_f <- phyloseq_filter_prevalence(ra_f, prev.trh=0.1, abund.trh=NULL)#filter out not present in <10% samples
asvs <- rownames(p_f@otu_table) #get names of ASVs
length(asvs) #get how many

#filter out phyloseq object
ps_Bac_f <- subset_taxa(ps_Bac_f, ASV %in% asvs)

#see how many reads were kept
sum(taxa_sums(ps_Bac_f))/sum(taxa_sums(ps_16S)) #59%
mean(taxa_sums(ps_Bac_f))
mean(sample_sums(ps_Bac_f))
#89% with final average of 3,808 reads per ASV and 71,508 reads per sample 

#clean
rm(asvs_to_keep, ra_ps, ra_f, p_f, asvs,t)

```

Now, normalize data with CSS transformation
```{r}

Bac_CSS <- lapply(Bac_exp_f, function(ps){phyloseq_transform_css(ps)})
lapply(Bac_CSS, function(ps){head(otu_table(ps))}) #ok

#same for single ps
ps_Bac_css <- phyloseq_transform_css(ps_Bac_f)
head(otu_table(ps_Bac_css))#ok

```

##Distances
```{r}

#calculate distance matrix, using 2 different indexes

#First with full ps

#Bray-Curtis
bray_full <- phyloseq::distance(ps_Bac_css, "bray")

#Unifrac
uni_full <- phyloseq::distance(ps_Bac_css, "unifrac") #problem

# there is a problem because the roots of the trees were probably prunned in the filtering step
# and then it goes a bit crazy, solution is to re-root or resolve multicotomies 
# used ape::multi2di(tree) to solve it
#thank you!: https://github.com/joey711/phyloseq/issues/936

ps_Bac_css@phy_tree <- ape::multi2di(ps_Bac_css@phy_tree) #change it

#Run Unifrac again
uni_full <- phyloseq::distance(ps_Bac_css, "unifrac") #ok


#Now per experiment

#Bray-Curtis
bray <- lapply(Bac_CSS, function(ps){ phyloseq::distance(ps, "bray")})

# first fix tree
Bac_CSS <- lapply(Bac_CSS, function(ps){
  
  tree <- ps@phy_tree
  tree2 <- ape::multi2di(tree)
  ps@phy_tree <- tree2
  return(ps)
}) #replace tree for the fixed one

#Now, run unifrac distances
uni <- lapply(Bac_CSS, function(ps){ phyloseq::distance(ps, "unifrac")}) #ok

```

##Ordination Plots

###Full
```{r}

dist <- list(bray=bray_full, uni=uni_full) #make it a list

#First get a short version of sample data df 
df_info_16S <- sample_data(ps_16S) %>% 
               data.frame() %>% 
               dplyr::select(Experiment, plant_pathway_induced)%>%
               mutate(sample_name=rownames(.))%>%
               filter(Experiment %in% c("cond_real_insects", "cond_phytohormones"))

#call function
get_UMAP_coordenates <- function(x){
  
  #dist is a list of distances previously calculated
  set.seed(123)
  umap <- umap(as.matrix(x))
  umap_scores <- as.data.frame(umap$layout)
  colnames(umap_scores) <- c("UMAP1", "UMAP2")
  umap_scores$sample_name <- rownames(umap_scores)
  umap_scores <- umap_scores %>% relocate(sample_name)
  return(umap_scores)
    
}

#get coordinates
df_UMAP <- lapply(dist, function(x){ get_UMAP_coordenates(x)})

#Now merge it with df 
df_UMAP <- lapply(df_UMAP, function(df){
  x <- left_join(df, df_info_16S, by="sample_name")
  x <- x %>% relocate(sample_name, Experiment, plant_pathway_induced)
})


#call function
plot_UMAP <- function(x){
  #x is df 
  #title is the name of the title in between ""
  set.seed(123)
  p <- ggplot(x, aes(x=UMAP1,y=UMAP2, color=plant_pathway_induced)) +
  geom_point(aes(color=plant_pathway_induced, shape=Experiment),
             size=4, alpha=0.85)+ # individual sample values
  #coord_fixed()+ # same axis scaling
  scale_color_manual(values=color_stress)+
  #ggtitle(title)+
  axis_looks+
  theme(legend.position="none")
  
  return(p)
}

#Plot it
#u.Bray <- plot_UMAP(df_UMAP$bray, "Bray")
u.uni <- plot_UMAP(df_UMAP$uni)

#Make it a single figure 
#ggarrange(u.Bray,u.uni, ncol=2, nrow=1,common.legend = TRUE, legend="bottom")

#save plot
mapply(function(x)
  ggsave(
  plot = last_plot(),
  filename =x,
  path = "./Plots/02_beta_diversity",
  scale = 1.5,
  width = 800,
  height = 750,
  units = "px",
  dpi = 300),
x=c("UMAP_Phyto-Insects_16S.svg","UMAP_Phyto-Insects_16S.png"))

```


###Phyto
Focus for now in nMDS and UMAP for Phyto
```{r}

phyto <- Bac_CSS$cond_phytohormones
dist <- list(bray=bray$cond_phytohormones, uni=uni$cond_phytohormones)

#NMDS

#call function
get_NMDS_coordenates <- function(x,y){
  
  #x is a phyloseq
  #y is a list of distances
  
  df_nmds <- lapply(y, function(dist){
    ord <- ordinate(x, "NMDS", distance=dist)
    nMDS <- plot_ordination(x, ord, color="plant_pathway_induced")
    df_NMDS <- nMDS$data
    df_NMDS <- df_NMDS %>% dplyr::select(Experiment, plant_pathway_induced, NMDS1, NMDS2)
    df_NMDS$sample_name <- rownames(df_NMDS)
    df_NMDS <- df_NMDS %>% relocate(sample_name)
    return(df_NMDS)
    })

  #ord will be a list 
  return(df_nmds)
}

#Get coordenates 
df_NMDS <- get_NMDS_coordenates(phyto,dist)

#Now, get centroids 
centroids <- lapply(df_NMDS, function(df){
  aggregate(cbind(NMDS1, NMDS2) ~ plant_pathway_induced, data=df, FUN=mean)})

#Finally, get segments [I can't make it into a function]
s.bray <- merge(df_NMDS$bray,centroids$bray, by.x="plant_pathway_induced", by.y="plant_pathway_induced",
        suffixes=c("","_centroid"))
s.uni <- merge(df_NMDS$uni,centroids$uni, by.x="plant_pathway_induced", by.y="plant_pathway_induced",
        suffixes=c("","_centroid"))

segments <- list(bray=s.bray, uni=s.uni)
rm(s.bray, s.uni)#clean

#Now, plot the NMDS
plot_NMDS <- function(x,y,z,title){
  #x is df 
  #y is the centroid df
  #z is segment df
  #title is the name of the title in between ""
  
  p <- ggplot(x, aes(x=NMDS1,y=NMDS2, color=plant_pathway_induced)) +
  geom_segment(data = z,
               mapping = aes(x=NMDS1, y=NMDS2, 
                             xend = NMDS1_centroid, yend = NMDS2_centroid),
               linewidth=0.3) + # spiders
  geom_point(size=0.9)+ # individual sample values
  geom_point(data = y, size = 5, alpha=0.9) + # centroids
  #coord_fixed()+ # same axis scaling
  scale_color_manual(values=color_stress)+
  ggtitle(title)
  return(p)
}

#Plot it
p.Bray <- plot_NMDS(df_NMDS$bray, centroids$bray, segments$bray, "Bray")
p.uni <- plot_NMDS(df_NMDS$uni, centroids$uni, segments$uni, "Unifrac")

#Make it a single figure 
ggarrange(p.Bray,p.uni, ncol=2, nrow=1,common.legend = TRUE,legend="bottom")

#save plot
mapply(function(x)
  ggsave(
  plot = last_plot(),
  filename =x,
  path = "./Plots/02_beta_diversity",
  scale = 1.8,
  width = 1000,
  height = 600,
  units = "px",
  dpi = 300),
x=c("NMDS_phyto_16S.svg","NMDS_phyto_16S.png"))

```

Now, do the same but with UMAP 

```{r}

#First get a short version of sample data df 
df_info_16S <- sample_data(ps_16S) %>% 
               data.frame() %>% 
               dplyr::select(Experiment, plant_pathway_induced)%>%
               mutate(sample_name=rownames(.))%>%
               filter(Experiment %in% c("cond_real_insects", "cond_phytohormones"))

#get coordinates
df_UMAP <- lapply(dist, function(x){ get_UMAP_coordenates(x)})

#Now merge it with df 
df_UMAP <- lapply(df_UMAP, function(df){
  x <- left_join(df, df_info_16S, by="sample_name")
  x <- x %>% relocate(sample_name, Experiment, plant_pathway_induced)
})

#calculate centroids 
centroids_UMAP <- lapply(df_UMAP, function(df){
  aggregate(cbind(UMAP1, UMAP2) ~ plant_pathway_induced, data=df, FUN=mean)})

#calculate segments
s.bray <- merge(df_UMAP$bray,centroids_UMAP$bray, by.x="plant_pathway_induced", by.y="plant_pathway_induced",
        suffixes=c("","_centroid"))
s.uni <- merge(df_UMAP$uni,centroids_UMAP$uni, by.x="plant_pathway_induced", by.y="plant_pathway_induced",
        suffixes=c("","_centroid"))

segments_UMAP <- list(bray=s.bray, uni=s.uni)
rm(s.bray, s.uni)#clean

plot_UMAP <- function(x,y,z,title){
  #x is df 
  #y is the centroid df
  #z is segment df
  #title is the name of the title in between ""
  
  p <- ggplot(x, aes(x=UMAP1,y=UMAP2, color=plant_pathway_induced)) +
  geom_segment(data = z, # spiders
               mapping = aes(x=UMAP1, y=UMAP2, 
                             xend = UMAP1_centroid, yend = UMAP2_centroid),
               linewidth=0.3) + 
  geom_point(size=1.5)+ # individual sample values
  geom_point(data = y, size = 6, alpha=0.9) + # centroids
  #coord_fixed()+ # same axis scaling
  scale_color_manual(values=color_stress)+
    #ggtitle(title)+
  axis_looks+
  theme(legend.position="none")
  return(p)
}

#Plot it
#u.Bray <- plot_UMAP(df_UMAP$bray, centroids_UMAP$bray, segments_UMAP$bray, "Bray")
u.uni <- plot_UMAP(df_UMAP$uni, centroids_UMAP$uni, segments_UMAP$uni, "Unifrac")

#Make it a single figure 
#ggarrange(u.Bray,u.uni, ncol=2, nrow=1,common.legend = TRUE, legend="bottom")

#save plot
mapply(function(x)
  ggsave(
  plot = last_plot(),
  filename =x,
  path = "./Plots/02_beta_diversity",
  scale = 1.5,
  width = 780,
  height = 750,
  units = "px",
  dpi = 300),
x=c("UMAP_phyto_16S.svg","UMAP_phyto_16S.png"))

#Finally, make a single df for phyto
#makes it a single df adding a column named "distance" with the name of the list
umap_phyto_df <- map_df(segments_UMAP, ~as.data.frame(.x), .id="distance") 
nmds_phyto_df <- map_df(segments, ~as.data.frame(.x), .id="distance") 

```

###Insects
Continue with Insects
```{r}

insects <- Bac_CSS$cond_real_insects
dist <- list(bray=bray$cond_real_insects, uni=uni$cond_real_insects)

#NMDS
#Get coordenates 
df_NMDS <- get_NMDS_coordenates(insects,dist)

#Now, get centroids 
centroids <- lapply(df_NMDS, function(df){
  aggregate(cbind(NMDS1, NMDS2) ~ plant_pathway_induced, data=df, FUN=mean)})

#Finally, get segments [I can't make it into a function]
s.bray <- merge(df_NMDS$bray,centroids$bray, by.x="plant_pathway_induced", by.y="plant_pathway_induced",
        suffixes=c("","_centroid"))
s.uni <- merge(df_NMDS$uni,centroids$uni, by.x="plant_pathway_induced", by.y="plant_pathway_induced",
        suffixes=c("","_centroid"))

segments <- list(bray=s.bray,uni=s.uni)
rm(s.bray, s.uni)#clean

#Now, plot the NMDS
p.Bray <- plot_NMDS(df_NMDS$bray, centroids$bray, segments$bray, "Bray")
p.uni <- plot_NMDS(df_NMDS$uni, centroids$uni, segments$uni, "Unifrac")

#Make it a single figure 
ggarrange(p.Bray,p.uni, ncol=2, nrow=1,common.legend = TRUE,legend="bottom")

#save plot
mapply(function(x)
  ggsave(
  plot = last_plot(),
  filename =x,
  path = "./Plots/02_beta_diversity",
  scale = 1.8,
  width = 1000,
  height = 600,
  units = "px",
  dpi = 300),
x=c("NMDS_insects_16S.svg","NMDS_insects_16S.png"))

```

Continue with UMAP

```{r}

#get coordinates
df_UMAP <- lapply(dist, function(x){ get_UMAP_coordenates(x)})

#Now merge it with df 
df_UMAP <- lapply(df_UMAP, function(df){
  x <- left_join(df, df_info_16S, by="sample_name")
  x <- x %>% relocate(sample_name, Experiment, plant_pathway_induced)
})

#calculate centroids 
centroids_UMAP <- lapply(df_UMAP, function(df){
  aggregate(cbind(UMAP1, UMAP2) ~ plant_pathway_induced, data=df, FUN=mean)})

#calculate segments
s.bray <- merge(df_UMAP$bray,centroids_UMAP$bray, by.x="plant_pathway_induced", by.y="plant_pathway_induced",
        suffixes=c("","_centroid"))
s.uni <- merge(df_UMAP$uni,centroids_UMAP$uni, by.x="plant_pathway_induced", by.y="plant_pathway_induced",
        suffixes=c("","_centroid"))

segments_UMAP <- list(bray=s.bray, uni=s.uni)
rm(s.bray, s.uni)#clean

#Now, plot the UMAP
#u.Bray <- plot_UMAP(df_UMAP$bray, centroids_UMAP$bray, segments_UMAP$bray, "Bray")
u.uni <- plot_UMAP(df_UMAP$uni, centroids_UMAP$uni, segments_UMAP$uni, "Unifrac")

#Make it a single figure 
#ggarrange(u.Bray,u.uni, ncol=2, nrow=1,common.legend = TRUE,legend="bottom")

#save plot
mapply(function(x)
  ggsave(
  plot = last_plot(),
  filename =x,
  path = "./Plots/02_beta_diversity",
  scale = 1.5,
  width = 780,
  height = 750,
  units = "px",
  dpi = 300),
x=c("UMAP_insects_16S.svg","UMAP_insects_16S.png"))

#Finally, make a single df for phyto
#makes it a single df adding a column named "distance" with the name of the list
umap_insects_df <- map_df(segments_UMAP, ~as.data.frame(.x), .id="distance") 
nmds_insects_df <- map_df(segments, ~as.data.frame(.x), .id="distance") 

```

Finally, let's place everything in a single df to save it 
```{r}

#makes it a single df adding a column named "distance" with the name of the list
umap_df <- rbind(umap_phyto_df,umap_insects_df )
nmds_df <- rbind(nmds_phyto_df,nmds_insects_df )

#save it as .csv df
write.csv(umap_df, file="./Data_frames/02_beta_diversity/umap_16S.csv")
write.csv(nmds_df, file="./Data_frames/02_beta_diversity/nmds_16S.csv")

#clean a little
rm(u.Bray,u.uni, p.Bray, p.uni, )#remove plots
rm(centroids, centroids_UMAP, segments, segments_UMAP, df_UMAP, df_NMDS)#df's

```

##PERMANOVA
```{r}

#I need this to test both experiments together 
ps_Bac_css
df_info_16S

#Is there a difference between the fungal communities among types of induction(Experiment) and treatments?
set.seed(123)

perm_full_bray <- adonis2(formula=distance(ps_Bac_css, "bray") 
                          ~ Experiment + plant_pathway_induced,
                     data=as(sample_data(ps_Bac_css), "data.frame"),
                     permutations = 999, by="margin") #no
set.seed(234)
perm_full_uni <- adonis2(formula=distance(ps_Bac_css, "unifrac") 
                          ~ Experiment + plant_pathway_induced,
                     data=as(sample_data(ps_Bac_css), "data.frame"),
                     permutations = 999, by="margin") #no, but barely

#No pairwise post-hoc

```
Save results
```{r}

#Permanova
perm_full_bray$distance <- "Bray-Curtis"
perm_full_bray <- perm_full_bray %>% rownames_to_column(var="factor") %>% relocate(distance)

perm_full_uni$distance <- "Unifrac"
perm_full_uni <- perm_full_uni %>% rownames_to_column(var="factor") %>% relocate(distance)

permanova_full <- rbind(perm_full_bray, perm_full_uni)

permanova_full <- permanova_full %>% mutate(across(where(is.numeric), ~ round(., 3)))


#save it
table <- nice_table(permanova_full,
                    highlight=TRUE,
                    title = c("Table X. PERMANOVA for the full 16S community"))

flextable::save_as_docx(table, path = "./Tables/02_beta_diversity/full_permanova_16S_BothExperiments.docx")

```


```{r}
#I need this to test both experiments independently
Bac_CSS
bray
uni

#Is there a difference between the fungal communities among treatments?
set.seed(123)

#Bray-Curtis
perm_Byexp_bray <- lapply(Bac_CSS, function(ps){
          set.seed(123)
          perm <- adonis2(formula=phyloseq::distance(ps, "bray") 
                          ~ plant_pathway_induced,
                     data=as(sample_data(ps), "data.frame"),
                     permutations = 999, by="margin")
          perm <- perm %>% rownames_to_column("factor")
          return(perm)
          })

perm_Byexp_bray <- map_df(perm_Byexp_bray, ~as.data.frame(.x), .id="Experiment") 
perm_Byexp_bray <- perm_Byexp_bray %>% mutate(distance="Bray-Curtis") %>% relocate(distance)

#Unifrac
perm_Byexp_uni <- lapply(Bac_CSS, function(ps){
          set.seed(123)
          perm <- adonis2(formula=phyloseq::distance(ps, "unifrac") 
                          ~ plant_pathway_induced,
                     data=as(sample_data(ps), "data.frame"),
                     permutations = 999, by="margin")
          perm <- perm %>% rownames_to_column("factor")
          return(perm)
          })

perm_Byexp_uni <- map_df(perm_Byexp_uni, ~as.data.frame(.x), .id="Experiment") 
perm_Byexp_uni <- perm_Byexp_uni %>% mutate(distance="Unifrac") %>% relocate(distance)

#merge
permanova_Byexp <- rbind(perm_Byexp_bray, perm_Byexp_uni)

#save it
table <- nice_table(permanova_Byexp,
              title = c("Table X. PERMANOVA test by Experiment for the 16S community"))
#flextable::save_as_docx(table, path = "./Tables/02_beta_diversity/permanova_ByExperiment_16S.docx")

#save df just in case 
write.table(permanova_Byexp, "./Data_frames/02_beta_diversity/permanova_byexperiment_16S.tsv", quote = F, row.names = F, sep = "\t")
```

```{r}
#save R enviroment
save.image(file="./R_Environments/02_beta_diversity/16S.Rdata")

```



```{r}
#Now, get pairwise differences but just for insects experiment, unifrac
#which treatments are different from each other?
pairwise_bray <- adonis.pair(dist.mat= phyloseq::distance(Bac_CSS$cond_real_insects, "bray"),
                 Factor =as.factor(as(sample_data(Bac_CSS$cond_real_insects),
                                      "data.frame")$plant_pathway_induced),
                 nper=1000,corr.method="fdr")#Ctrl vs JA

pairwise_uni <- adonis.pair(dist.mat= phyloseq::distance(Bac_CSS$cond_real_insects, "unifrac"),
                 Factor = as.factor(as(sample_data(Bac_CSS$cond_real_insects),
                                       "data.frame")$plant_pathway_induced),
                 nper=1000,corr.method="fdr")#Ctrl vs JA

pairwise_bray$distance <- "Bray-Curtis"
pairwise_bray <- pairwise_bray %>% relocate(distance)

pairwise_uni$distance <- "Unifrac"
pairwise_uni <- pairwise_uni %>% relocate(distance)

pairwise_full <- rbind(pairwise_bray, pairwise_uni) #merge it

#save it
table <- nice_table(pairwise_full ,
              title = c("Table X. Overall pairwise comparisons between treatments of the 16S community for the Insect Herbivores experiment "))
flextable::save_as_docx(table, path = "./Tables/02_beta_diversity/phyto_pairwise_ITS.docx")

```

#Summary
Now, here's my attempt of a written summary:

For fungi samples that were below 195,000 and above 9,500 reads were kept which corresponds to roughly 1x the standard deviation from the mean. For Bacteria, samples that were below 200,000 and above 10,000 reads were kept which corresponded to roughly 2x the standard deviation from the mean.For fungi this resulted in the removal of 22 out of 145 samples in total for both experiments whereas for bacteria filtering resulted in 13 out of 146 samples in total for both experiments. For beta-diversity and differential abundance analysis ASVs were further filtered independently by experiment by keeping those with relative abundance higher than 0.001% and present in at least 10% of the samples. This resulted in an average of 98,849 reads/sample for fungi and 71,956 reads/sample for bacteria. The final number of unique ASVs was 167 for fungi and 2,464 for bacteria, on average by experiment (Supplementary table XX). 


For fungi, treatment had an effect on the composition of the community when using Bray-Curtis distances (PERMANOVA, XX,XX,XX) but only for the phytohormone-induced experiment. The same result was observed for UNIFRAC distances. Particularly, control samples were different than those belonging to JA-induced plants (result, p value). For Bacteria, there was no major effect of the treatment in the beta-diversity when measured by Bray-Curtis distances nor by UNIFRAC (Supp. Figure X). 

*I still need to make the filtering by experiment
*run alpha diversity analysis at Family level (perhaps more informative)
*Make a ternary plot (perhaps also at family level)





