---
title: "02_phytohormones.Rmd"
author: "Marcela Aragon"
date: "10/1/2024"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: 
     collapsed: false
     smooth_scroll: true
    df_print: paged
    code_folding: hide
    keep_md: true
editor_options: 
  chunk_output_type: console
---

This script will deal with ploting and analyzing the concentration of X different phytohormones in both conditioning and feedback phases;

# General settings 

## Rmarkdown

```{r global settings, include=FALSE}

#Set global options for r chunks
knitr::opts_chunk$set(
  echo = TRUE,#show code
  fig.width = 9,
  fig.height = 6,
  dpi = 300,
  error=TRUE, #keep knitting even if error appears
  message=FALSE, #don't show messages
  warning=FALSE, #don't show warnings
  collapse=FALSE)

```

## Loading packages

```{r loading packages, echo=TRUE}

#Set working directory to project directory 
setwd("./")
getwd() #ok

#data
library(knitr)
library(dplyr)
library(tibble)

#plotting
library(ggplot2)
library(ggpubr)
library(GGally)
library(RColorBrewer)
library(grDevices)
library(rempsyc) #tables

#stats
library(DHARMa)
library(lmtest)
library(fitdistrplus)
library(AICcmodavg)
library(emmeans)
library(multcompView)
library(multcomp)
library(postHoc)
library(car)
library(effsize) #Cohen's D
library(lsr)

library(ropls) #PSLDA
library(PERMANOVA)
library(plsdepot)
library(santaR)
library(vegan)


```

## Global settings for plots 

```{r settings for plots}

# Setting up white background for plots
theme_set(theme_bw())

# Set colors
color_JA <- rgb(0.9019607843137255, 0.6235294117647059, 0) #yellow-orange
color_SA <- rgb(0.26666666666666666,0.4470588235294118,0.7686274509803922) #blue
color_control <- rgb(0.6,0.6,0.6) #gray
color_simulated <- rgb(0.8,0.2,0.6) #pink
color_real_insects <- rgb(0, 0.6196078431372549, 0.45098039215686275) #green
color_non-conditioned <-rgb(1,1,1,1)

color_pathway_induced <- c(color_control, color_JA, color_SA) #control, JA and SA
color_experiment <- c(color_simulated, color_real_insects) #simulated and real herbivory

# Set default axis looks
axis_looks <- theme(axis.text.x = element_text(colour = "black", size = 9,
                                               face = "bold", angle=0, hjust=0.5))+ #text in X
  theme(axis.text.y = element_text(colour = "black", size = 9, face = "bold"))+ #text in y
  theme(axis.title=element_text(size=11, face = "bold"))+ #all texts
  theme(axis.title.y = element_text(margin = ggplot2::margin(t = 0, r = 10, b = 0, l = 0)))+
  theme(panel.grid.major = element_blank(), #no grid
        panel.grid.minor = element_blank()) #no grid
  #theme(legend.position="none")+ #no legend

#Get names for boxes in facet wrap
Experiment_names <- c(
                    `simulated` = "Simulated Herbivory",
                    `real_insects` = "Insect Herbivory")

```

## Conditioning 

The conditioning phase for the simulated-herbivory experiment took place as part of the Family Experiment at Wageninen in between May and June from 2021. And the insect-herbivory experiment took place between *X and X from 2021.*

From the 16-17 total replicates in the conditioning phase, 4 plants per treatment were randomly allocated to be used for metabolite analysis. For this, shoot biomass was cut from the root and flash-froze in liquid nitrogen. Samples were freeze-dried for 72 hours (*equipment*) and grinded to fine powder at the UNIFARM facilities in Wageningen (big extruder in grinding room). Per sample, 15-30 mg (*I need to check this*) were weighed individually and placed in a special tray (*get the name*) to be quantified by LC-MS/MS in Jenna by Michael Reichelt.

Text by Michael Reichelt: 

"Phytohormone analysis was performed as in Vadassery et.al. (2012) on an Agilent 1200 series HPLC system (Agilent Technologies) with the modification that a tandem mass spectrometer QTRAP 6500 (SCIEX, Darmstadt, Germany) was used. Since it was observed that  both  the  D6-labeled  JA  and  D6-labeled  JA-Ile  standards (HPC Standards GmbH, Cunnersdorf, Germany) contained 40% of the corresponding D5-labeled compounds, the sum of the peak areas of D5- and D6-compound was used for quantification. Concentration of cis-OPDA, and OH-JA were determined relative to the quantity of the internal standard D6-JA applying a response factor (RF) of 1.0. OH-JA-Ile and COOH-JA-Ile were quantified relative to  D6-JA-Ile: RF 1.0. Sulfo-JA was determined relative to the quantity of the internal standard D6-JA: RF 6.0.

Data processing was performed using Analyst 1.5 software and analyte quantity was determined relative to the corresponding internal standard peak area."  

### Exploring data 


```{r}

#load data
phyto <- readRDS("./plant_data/phytohormones.rds")

#explore 
summary(phyto)
colnames(phyto)

#change levels of Experiment
levels(phyto$Experiment)
phyto <- phyto %>%
                mutate(Experiment = factor(Experiment,
                                           levels = c("simulated", "real_insects" )))%>%
                mutate(plant_pathway_induced = factor(plant_pathway_induced,
                                                      levels=c("Control", "JA", "SA")))

#subsetting df to only conditioning
conditioning <- phyto %>% 
                filter(Phase == "conditioning")

#make long format for plotting
long <- phyto %>% 
              tidyr::pivot_longer(cols = c(11:18), 
              names_to = "phytohormone",
              values_to = "concentration",
              values_drop_na = TRUE)

#making new phytohormone column as factor
long$phytohormone <- as.factor(long$phytohormone)

# Reordering downstream the JA pathway
ordered_hormones <- c("ABA", "SA", "cis_OPDA", "JA", "OH_JA", "sum_JA_Ile", "OH_JA_Ile", "COOH_JA_Ile")

long$phytohormone <-factor(long$phytohormone, 
                             levels = ordered_hormones)

#subsetting df to only conditioning
conditioning_long <- long %>% 
                filter(Phase == "conditioning")


#plotting quickly
p <- ggpairs(phyto[c("Experiment","Phase","plant_pathway_induced", "JA", "SA")], 
            mapping = aes(color = Experiment))

#adding colors
p + scale_color_manual(values=color_experiment)+
           scale_fill_manual(values=color_experiment)

#only conditioning
p <- ggpairs(conditioning[c("Experiment","plant_pathway_induced", "phytohormone", "concentration")], 
            mapping = aes(color = Experiment))

#adding colors
p + scale_color_manual(values=color_experiment)+
           scale_fill_manual(values=color_experiment)

#clean
rm(p)
dev.off()

```

Now I have 3 df, phyto, long and conditioning 


* Values look similar between experiments in the conditioning phase
* There are slightly less replicates for SA

### Plots

#### Barplots

Boxplot with rawdata points, colored by plant_pathway_induced and facet wraped by Experiment + colored in the upper boxes. For this, first I need to make a summary table df. 

```{r}
#summary stats 
df_phyto <- conditioning_long %>%
  group_by(Experiment, phytohormone, plant_pathway_induced) %>%
  summarise(
    n = n(),
    mean = mean(concentration, na.rm = TRUE),
    sd = sd(concentration, na.rm = TRUE)) %>%
    mutate(se = sd/sqrt(n))

df_phyto

```

Now, let's plot it. I'd like to have a plot with those hormones with high concentrations and then a second plot with a smaller axis for those of low concentrations so differences can be better appreciated.

```{r}

big_axis <- c("ABA", "SA", "cis_OPDA", "JA", "OH_JA")
small_axis <- c("sum_JA_Ile", "OH_JA_Ile","COOH_JA_Ile")


# make a barplot for big-axis
p1 <- ggplot(data = subset(df_phyto, phytohormone %in% big_axis),
            mapping = aes(x = phytohormone, y = mean), colour= plant_pathway_induced)+
  geom_bar(aes(fill = plant_pathway_induced), stat = "identity",
           position = position_dodge(0.8), width = 0.8, color = "black")+
  geom_errorbar(aes(ymin = mean-se, ymax = mean+se, group = plant_pathway_induced),
                width = 0.2, position = position_dodge(0.8))+
  scale_fill_manual(values = color_pathway_induced)+
  geom_point(data = subset(conditioning_long, phytohormone %in% big_axis),
              mapping = (aes(phytohormone, concentration, fill = plant_pathway_induced)),
               shape = 21, size = 1.5, color = "black",
              position = position_jitterdodge(jitter.height = .1, jitter.width = .2))

p2 <- p1 + facet_wrap(~Experiment, labeller = as_labeller(Experiment_names)) +
              labs(y="(ng/g DW)", x=" ") +
              axis_looks +
              theme(panel.spacing = unit(0.2, "lines"))+
              theme(legend.position = "none")+
              scale_x_discrete(labels=c("ABA" = "ABA", "SA" = "SA",
                                      "cis_OPDA" = "cis-OPDA", "JA" = "JA", "OH_JA" = "OH-JA"))
#save plot
ggsave(
  filename = "cond_phyto_big.svg",
  plot = p2,
  path = "./Plots/02_phytohormones",
  scale = 1.5,
  width = 1200,
  height = 700,
  units = "px",
  dpi = 300)


# make a barplot for small-axis
p1 <- ggplot(data = subset(df_phyto, phytohormone %in% small_axis),
            mapping = aes(x = phytohormone, y = mean), colour= plant_pathway_induced)+
  geom_bar(aes(fill = plant_pathway_induced), stat = "identity",
           position = position_dodge(0.8), width = 0.8, color = "black")+
  geom_errorbar(aes(ymin = mean-se, ymax = mean+se, group = plant_pathway_induced),
                width = 0.2, position = position_dodge(0.8))+
  scale_fill_manual(values = color_pathway_induced)+
  geom_point(data = subset(conditioning_long, phytohormone %in% small_axis),
              mapping = (aes(phytohormone, concentration, fill = plant_pathway_induced)),
               shape = 21, size = 1.5, color = "black",
              position = position_jitterdodge(jitter.height = .1, jitter.width = .2))

p2 <- p1 + facet_wrap(~Experiment) +
            theme(strip.background = element_blank(),
                  strip.text.x = element_blank())+ #no label on top
              labs(y="(ng/g DW)", x=" ") +
              axis_looks +
              theme(panel.spacing = unit(0.2, "lines"))+
              theme(legend.position = "none")+
              scale_x_discrete(labels=c("sum_JA_Ile" = "JA-Ile", "OH_JA_Ile" = "OH-JA-Ile",
                                      "COOH_JA_Ile" = "COOH-JA-Ile"))

#save plot
ggsave(
  filename = "cond_phyto_small.svg",
  plot = p2,
  path = "./Plots/02_phytohormones",
  scale = 1.5,
  width = 1200,
  height = 700,
  units = "px",
  dpi = 300)

#clean
rm(p1,p2)
dev.off()

```

These two plots will be fused in one in illustrator


#### PLSDA

Stands for Partial Least Squares Discriminant Analysis; a type of multivariate dimensionality reduction analysis. It can be thought as a "supervised" version of a PCA 


```{r}

#The following code was written by Els van de Zande, thank you Els!

View(conditioning)

str(conditioning)
matrixdata<-conditioning[,11:18]# keep only values
matrixdata <- as.matrix(matrixdata[,])# make it numeric
metadata<-conditioning[,c("Experiment", "plant_pathway_induced")]
metadata <- metadata %>% #change from characters to numbers
              mutate(Experiment = case_when(Experiment == 'simulated' ~ '1',
                                            TRUE ~ '2')) %>%
              mutate(plant_pathway_induced = case_when(plant_pathway_induced == 'Control' ~ '1',
                                                       plant_pathway_induced == 'JA' ~ '2',
                                                       TRUE ~ '3'))
MetaData <-data.matrix(metadata) #makes a numeric matrix

#PLS-DA
plsda <-opls(matrixdata,MetaData)
plsda_center <-opls(matrixdata,MetaData, scaleC = "center")
plsda_pareto <-opls(matrixdata,MetaData, scaleC = "pareto")#removes the relative importance of large values
                                                          #but keeps the data structure intact.
#get scores to plot them
scores<-plsda_pareto@scoreMN
scores_df<-as.data.frame(scores) #make new df 

scores_df$p1 <- as.numeric(scores_df$p1)
scores_df$p2 <- as.numeric(scores_df$p2)

scores_df2 <- scores_df %>% #add info
              mutate(Experiment = as.numeric(metadata$Experiment))%>%
              mutate(plant_pathway_induced = as.numeric(metadata$plant_pathway_induced))
              

#Plot the scores using ggplot2
scoreplot<-ggplot(scores_df2, aes(x = p1, y = p2, fill = factor(plant_pathway_induced),
                                  shape = factor(Experiment))) +
  geom_point(aes(colour=factor(plant_pathway_induced), fill=factor(plant_pathway_induced)),
              size = 4, alpha=0.8)+
  labs(x = "axis 1", y = "axis 2",
       color="Plant pathway induced", shape="Experiment", fill=element_blank()) +
  scale_fill_manual(values =color_pathway_induced) +  #Customize color scheme for Treatment
  scale_color_manual(values =color_pathway_induced)
  scale_shape_manual(values = c(16,17)) +  #Customize shape for Herbivore
  axis_looks

#now add loadings
loadings<-plsda_pareto@loadingMN
loadings_df<-as.data.frame(loadings)

loadings_df$p1 <- as.numeric(loadings_df$p1)
loadings_df$p2 <- as.numeric(loadings_df$p2)

#makes plot with arrows marking direction
hormone_names <- c("SA","JA","ABA","JA-Ile","cis-OPDA","OH-JA","OH-JA-Ile","COOH-JA-Ile")

Biplot <-scoreplot+
         geom_hline(yintercept=0, linetype='dotted', col = 'gray20',linewidth = 0.2 )+
         geom_vline(xintercept=0, linetype='dotted', col = 'gray20', linewidth = 0.2)+
         geom_segment(data = loadings_df, aes(x = 0, y = 0, xend = (p1*50), yend = (p2*50)),
                      arrow = arrow(length = unit(0.5, "picas"), type="closed"),
                      colour = "black", linewidth = 0.5, inherit.aes = FALSE)+
         geom_text(aes(x = (p1*50), y = (p2*50), label = hormone_names,
                angle = 0.45, hjust = -.1, vjust = .2),
                 data = loadings_df,
                 colour = "black", size=4, inherit.aes = FALSE)+
         axis_looks
         #theme(legend.position = "none")

#save plot
ggsave(
  filename = "cond_phyto_PLSDA.svg",
  plot = Biplot,
  path = "./Plots/02_phytohormones",
  scale = 1.5,
  width = 1200,
  height = 700,
  units = "px",
  dpi = 300)


#extract VIP (Variable's importance) scores and CV-value
VIPmatrixdataPLSDA<-getVipVn(plsda_pareto, orthoL = FALSE)
write.csv(VIPmatrixdataPLSDA)

#PERMANOVA
UVmatrixdata<- santaR:::scaling_UV(matrixdata) #for auto-scaling using st. deviation
DistanceMatrix<-dist(UVmatrixdata, method="euclidean")

#make it a df again
MetaData <-as.data.frame(MetaData) 
MetaData[,c(1:2)]<- lapply(MetaData[,c(1:2)], as.factor)

#run PERMANOVA
permanova <- adonis2(DistanceMatrix~Experiment*plant_pathway_induced,
                      data=MetaData, permutations = 999, method="bray")






```






### Stats

For now, let's analyze each phytohormone independently to see whether concentration changes according to the plant_pathway_induced. Thus, one GLM per hormone per experiment. Try to gather a final table with values from the model for everything. 

```{r}

#make 2 df, one per Experiment
simulated <- subset(conditioning, Experiment == "simulated") %>% droplevels()
real <- subset(conditioning, Experiment == "real_insects") %>% droplevels()

#make a list
list <- list(simulated = simulated, real=real)

#use lapply for each hormone
ABA <- lapply(list, function(df){
              glm(log(concentration) ~ plant_pathway_induced,
               family = Gamma (link = "log"),
               data = subset(df, phytohormone == "ABA"))})

#check
simulateResiduals(ABA$simulated, plot = T)
simulateResiduals(ABA$real, plot = T)

#get df 
summary(ABA$simulated)
summary(ABA$real)

```





