---
title: "02_phytohormones.Rmd"
author: "Marcela Aragon"
date: "10/1/2024"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: 
     collapsed: false
     smooth_scroll: true
    df_print: paged
    code_folding: hide
    keep_md: true
editor_options: 
  chunk_output_type: console
---

This script will deal with ploting and analyzing the concentration of X different phytohormones in both conditioning and feedback phases;

# General settings 

## Loading packages

```{r loading packages, echo=TRUE}

#Set working directory to project directory 
setwd("./")
getwd() #ok

#data
library(knitr)
library(dplyr)
library(tibble)

#plotting
library(ggplot2)
library(ggpubr)
library(GGally)
library(RColorBrewer)
library(grDevices)
library(rempsyc) #tables

#stats
library(DHARMa)
library(lmtest)
library(fitdistrplus)
library(AICcmodavg)
library(emmeans)
library(multcompView)
library(multcomp)
library(postHoc)
library(car)
library(effsize) #Cohen's D
library(lsr)

library(ropls) #PLS-DA
library(mixOmics) #PLS-DA
library(PERMANOVA)
library(plsdepot)
library(santaR)
library(vegan)


```

## Global settings for plots 

```{r settings for plots}

# Setting up white background for plots
theme_set(theme_bw())

#Set colors
color_stress <- c("#BFBFBF", "#E69F00", "#4472C4")
color_experiment <- c("#CC3399", "#70AD47")

color_ctrl_vs_JA <- c("#BFBFBF", "#E69F00")
color_ctrl_vs_SA <- c("#BFBFBF",  "#4472C4")
color_control <- "#BFBFBF"
color_JA <- "#E69F00"
color_SA <- "#4472C4"
color_experiment <- c("#CC3399", "#70AD47")
fonts <- windowsFonts()

color_pathway_induced <- c(color_control, color_JA, color_SA) #control, JA and SA


# Set default axis looks
axis_looks <- theme(axis.text.x = element_text(colour = "black", size = 9,
                                               face = "bold", angle=0, hjust=0.5))+ #text in X
  theme(axis.text.y = element_text(colour = "black", size = 9, face = "bold"))+ #text in y
  theme(axis.title=element_text(size=11, face = "bold"))+ #all texts
  theme(axis.title.y = element_text(margin = ggplot2::margin(t = 0, r = 10, b = 0, l = 0)))+
  theme(panel.grid.major = element_blank(), #no grid
        panel.grid.minor = element_blank()) #no grid
  #theme(legend.position="none")+ #no legend

#Get names for boxes in facet wrap
Experiment_names <- c(
                    `simulated` = "Phytohormone-treated",
                    `real_insects` = "Herbivore-infested")

```

## Conditioning 

The conditioning phase for the simulated-herbivory experiment took place as part of the Family Experiment at Wageninen in between May and June from 2021. And the insect-herbivory experiment took place between *X and X from 2021.*

From the 16-17 total replicates in the conditioning phase, 4 plants per treatment were randomly allocated to be used for metabolite analysis. For this, shoot biomass was cut from the root and flash-froze in liquid nitrogen. Samples were freeze-dried for 72 hours (*equipment*) and grinded to fine powder at the UNIFARM facilities in Wageningen (big extruder in grinding room). Per sample, 15-30 mg (*I need to check this*) were weighed individually and placed in a special tray (*get the name*) to be quantified by LC-MS/MS in Jenna by Michael Reichelt.

Text by Michael Reichelt: 

"Phytohormone analysis was performed as in Vadassery et.al. (2012) on an Agilent 1200 series HPLC system (Agilent Technologies) with the modification that a tandem mass spectrometer QTRAP 6500 (SCIEX, Darmstadt, Germany) was used. Since it was observed that  both  the  D6-labeled  JA  and  D6-labeled  JA-Ile  standards (HPC Standards GmbH, Cunnersdorf, Germany) contained 40% of the corresponding D5-labeled compounds, the sum of the peak areas of D5- and D6-compound was used for quantification. Concentration of cis-OPDA, and OH-JA were determined relative to the quantity of the internal standard D6-JA applying a response factor (RF) of 1.0. OH-JA-Ile and COOH-JA-Ile were quantified relative to  D6-JA-Ile: RF 1.0. Sulfo-JA was determined relative to the quantity of the internal standard D6-JA: RF 6.0.

Data processing was performed using Analyst 1.5 software and analyte quantity was determined relative to the corresponding internal standard peak area."  

### Exploring data 


```{r}

#load data
phyto <- readRDS("./plant_data/phytohormones.rds")

#explore 
summary(phyto)
colnames(phyto)

#change levels of Experiment
levels(phyto$Experiment)
levels(phyto$plant_pathway_induced)
phyto <- phyto %>%
                mutate(Experiment = factor(Experiment,
                                           levels = c("simulated", "real_insects" )))%>%
                mutate(plant_pathway_induced = factor(plant_pathway_induced,
                                                      levels=c("Control", "JA", "SA")))

#subsetting df to only conditioning
conditioning <- phyto %>% 
                filter(Phase == "conditioning")

#make long format for plotting
long <- phyto %>% 
              tidyr::pivot_longer(cols = c(11:18), 
              names_to = "phytohormone",
              values_to = "concentration",
              values_drop_na = TRUE)

#making new phytohormone column as factor
long$phytohormone <- as.factor(long$phytohormone)

# Reordering downstream the JA pathway
ordered_hormones <- c("ABA", "SA", "cis_OPDA", "JA", "OH_JA", "sum_JA_Ile", "OH_JA_Ile", "COOH_JA_Ile")

long$phytohormone <-factor(long$phytohormone, 
                             levels = ordered_hormones)

#subsetting df to only conditioning
conditioning_long <- long %>% 
                filter(Phase == "conditioning")

```

* Now I have 3 df, phyto, long and conditioning 

Plotting quickly
```{r}

#plotting quickly
#p <- ggpairs(phyto[c("Experiment","Phase","plant_pathway_induced", "JA", "SA")], 
#            mapping = aes(color = Experiment))

#adding colors
#p + scale_color_manual(values=color_experiment)+
#           scale_fill_manual(values=color_experiment)

#only conditioning
#p <- ggpairs(conditioning[c("Experiment","plant_pathway_induced", "phytohormone", "concentration")], 
#            mapping = aes(color = Experiment))

#adding colors
#p + scale_color_manual(values=color_experiment)+
#           scale_fill_manual(values=color_experiment)

#clean
#rm(p)
#dev.off()

```

* Values look similar between experiments in the conditioning phase
* There are slightly less replicates for SA

### PLS-DA 

Stands for Partial Least Squares Discriminant Analysis; a type of multivariate dimensionality reduction analysis. It can be thought as a "supervised" version of a PCA. The general approach is to analyze all phytohormones in a single analysis to check for patterns.  

https://omicsforum.ca/t/what-is-the-meaning-of-components-in-pls-da/732/3

```{r}

#Some of the following code was written by Els van de Zande, thank you Els!

View(conditioning)

str(conditioning)
matrixdata<-conditioning[,11:18]# keep only values
matrixdata <- as.matrix(matrixdata[,])# make it numeric
metadata<-conditioning[,c("Experiment", "plant_pathway_induced")]
metadata <- metadata %>% mutate(joined=paste(Experiment, plant_pathway_induced, sep="_"))
metadata <- factor(metadata$joined)

#PLS-DA 
plsda <-opls(matrixdata, metadata, scaleC = "pareto", permI = 100)
#pareto removes the relative importance of large values
#but keeps the data structure intact.

#get summary
plot(plsda, typeVc = "x-score", parCexN = 1.5)
vip_scores <- getVipVn(plsda)

#let's remove those with VIP < 0.75
low_vip <- which(vip_scores < 0.75)
filtered_matrixdata <- matrixdata[,-low_vip]

#re-run PLSDA
plsda_filt <-opls(filtered_matrixdata, metadata, scaleC = "pareto", permI = 100)
plot(plsda_filt)

#get scores
scores<-plsda_filt@scoreMN
scores_df<-as.data.frame(scores) #make new df 

scores_df$p1 <- as.numeric(scores_df$p1)
scores_df$p2 <- as.numeric(scores_df$p2)

metadata<-conditioning[,c("Experiment", "plant_pathway_induced")]

scores_df2 <- scores_df %>% #add info
              mutate(Experiment = as.numeric(metadata$Experiment))%>%
              mutate(plant_pathway_induced = as.numeric(metadata$plant_pathway_induced))

```

Old code 
```{r}

#split by factors
plant_pathway_Factor <- MetaData[,"plant_pathway_induced"]
experiment_Factor <- MetaData[,"Experiment"]

plsda_pareto2 <-opls(matrixdata,plant_pathway_Factor, scaleC = "pareto", permI=1000) #plant pathway induced
plsda_pareto3 <-opls(matrixdata,experiment_Factor, scaleC = "pareto", permI=1000) #Experiment 

#get scores to plot them
scores<-plsda_pareto@scoreMN
scores_df<-as.data.frame(scores) #make new df 

scores_df$p1 <- as.numeric(scores_df$p1)
scores_df$p2 <- as.numeric(scores_df$p2)

scores_df2 <- scores_df %>% #add info
              mutate(Experiment = metadata$Experiment)%>%
              mutate(plant_pathway_induced = metadata$plant_pathway_induced)
              
```

Biplot
```{r}
#Plot the scores using ggplot2
scoreplot<-ggplot(scores_df2, aes(x = p1, y = p2, fill = factor(plant_pathway_induced),
                                  shape = factor(Experiment))) +
  geom_point(aes(colour=factor(plant_pathway_induced), fill=factor(plant_pathway_induced)),
              size = 4, alpha=0.8)+
  labs(x = "t1(69%)", y = "t2(17%)",
       color="Plant pathway induced", shape="Experiment", fill=element_blank()) +
  scale_fill_manual(values =color_pathway_induced) +  #Customize color scheme for Treatment
  scale_color_manual(values =color_pathway_induced)
  scale_shape_manual(values = c(16,17)) +  #Customize shape for Herbivore
  axis_looks

#now add loadings
loadings<-plsda_filt@loadingMN
loadings_df<-as.data.frame(loadings)

loadings_df$p1 <- as.numeric(loadings_df$p1)
loadings_df$p2 <- as.numeric(loadings_df$p2)

#makes plot with arrows marking direction
hormone_names <- c("SA","JA","ABA","cis-OPDA","OH-JA")

Biplot <-scoreplot+
         geom_hline(yintercept=0, linetype='dotted', col = 'gray20',linewidth = 0.2 )+
         geom_vline(xintercept=0, linetype='dotted', col = 'gray20', linewidth = 0.2)+
         geom_segment(data = loadings_df, aes(x = 0, y = 0, xend = (p1*50), yend = (p2*50)),
                      arrow = arrow(length = unit(0.5, "picas"), type="closed"),
                      colour = "black", linewidth = 0.5, inherit.aes = FALSE)+
         geom_text(aes(x = (p1*50), y = (p2*50), label = hormone_names,
                angle = 0.45, hjust = -.1, vjust = .2),
                 data = loadings_df,
                 colour = "black", size=4, inherit.aes = FALSE)+
        #ylim(-60,60)+
        #xlim(-100,100)+
        axis_looks
         #theme(legend.position = "none")
Biplot

#save plot
mapply(function(x)
ggsave(
  plot = last_plot(),
  filename =x,
  path = "./Plots/02_phytohormones",
  scale = 1.5,
  width = 1200,
  height = 700,
  units = "px",
  dpi = 300),
x=c("cond_phyto_PLSDA_filtered.svg",
    "cond_phyto_PLSDA_filtered.png"))

```

Now, let's analyze the data with a PERMANOVA. First, data is auto-scaled to calculate a distance matrix from which a PERMANOVA is run to see whether there is an effect of plant pathway induced, the type of induction and the interaction of these two factors. 

### PERMANOVA

```{r}

#PERMANOVA
UVmatrixdata<- santaR:::scaling_UV(filtered_matrixdata) #for auto-scaling using st. deviation
DistanceMatrix<-dist(UVmatrixdata, method="euclidean")

#make it a df again
MetaData <-as.data.frame(metadata) 
MetaData[,c(1:2)]<- lapply(MetaData[,c(1:2)], as.factor)

#run PERMANOVA
permanova <- adonis2(DistanceMatrix~Experiment*plant_pathway_induced,
                      data=MetaData, permutations = 999, method="bray")

#make table for Word
. <- c("Experiment (E)", "Plant pathway induced (PPI)", "E*PPI", "Residual", "Total") #names of factors
permanova_table <- cbind (.,as.data.frame(permanova))

table <- nice_table(permanova_table,
              title = c("Table X. PERMANOVA for Plant Phytohormones", "Conditioning phase"),
              note = "999 number of permutations")

flextable::save_as_docx(table, path = "./Tables/02_phytohormones/permanova_phytohormones.docx")
```

```{r}
#Run post-hoc tests

#call pairwise.adonis2 function
pairwise.adonis2 <- function(x, factors, sim.method = "bray", p.adjust.m = "bonferroni") {
    co = combn(unique(factors), 2)
    pairs = c()
    F.Model = c()
    R2 = c()
    p.value = c()
    for (elem in 1:ncol(co)) {
        ad = adonis2(x[factors %in% c(co[1, elem], co[2, elem]), ] ~ factors[factors %in% c(co[1, elem], co[2, elem])],
                     permutations = 999, method = sim.method)
        pairs = c(pairs, paste(co[1, elem], 'vs', co[2, elem]))
        F.Model = c(F.Model, ad$F[1])
        R2 = c(R2, ad$R2[1])
        p.value = c(p.value, ad$`Pr(>F)`[1])
    }
    p.adjusted = p.adjust(p.value, method = p.adjust.m)
    pairwise_results = data.frame(pairs, F.Model, R2, p.value, p.adjusted)
    return(pairwise_results)
}

#run it
pairwise_results <- pairwise.adonis2(as.matrix(DistanceMatrix), MetaData$plant_pathway_induced)
pairwise_results$p.adjusted <- format(pairwise_results$p.adjusted, digits = 3, nsmall = 3) #to keep 3 decimals after the point

#save it
table <- nice_table(pairwise_results,
              title = c("Table X. Post-Hoc PERMANOVA for Plant Phytohormones", "Conditioning phase"))
              
flextable::save_as_docx(table, path = "./Tables/02_phytohormones/posthoc_cond_ppi.docx")

```


### Individual Barplots

Boxplot with rawdata points, colored by plant_pathway_induced and facet wraped by Experiment + colored in the upper boxes. For this, first I need to make a summary table df. 

```{r}
#summary stats 
df_phyto <- conditioning_long %>%
  group_by(Experiment, phytohormone, plant_pathway_induced) %>%
  summarise(
    n = n(),
    mean = mean(concentration, na.rm = TRUE),
    sd = sd(concentration, na.rm = TRUE)) %>%
    mutate(se = sd/sqrt(n))

df_phyto

```

Now, let's plot it. I'd like to have a plot with those hormones with high concentrations and then a second plot with a smaller axis for those of low concentrations so differences can be better appreciated.

```{r}

big_axis <- c("ABA", "SA", "cis_OPDA", "JA", "OH_JA")
small_axis <- c("sum_JA_Ile", "OH_JA_Ile","COOH_JA_Ile")


# make a barplot for big-axis
p1 <- ggplot(data = subset(df_phyto, phytohormone %in% big_axis),
            mapping = aes(x = phytohormone, y = mean), colour= plant_pathway_induced)+
  geom_bar(aes(fill = plant_pathway_induced), stat = "identity",
           position = position_dodge(0.8), width = 0.8, color = "black")+
  geom_errorbar(aes(ymin = mean-se, ymax = mean+se, group = plant_pathway_induced),
                width = 0.2, position = position_dodge(0.8))+
  scale_fill_manual(values = color_pathway_induced)+
  geom_point(data = subset(conditioning_long, phytohormone %in% big_axis),
              mapping = (aes(phytohormone, concentration, fill = plant_pathway_induced)),
               shape = 21, size = 1.5, color = "black",
              position = position_jitterdodge(jitter.height = .1, jitter.width = .2))

p2 <- p1 + facet_wrap(~Experiment, labeller = as_labeller(Experiment_names)) +
              labs(y="(ng/g DW)", x=" ") +
              axis_looks +
              theme(panel.spacing = unit(0.2, "lines"))+
              theme(legend.position = "none")+
              scale_x_discrete(labels=c("ABA" = "ABA", "SA" = "SA",
                                      "cis_OPDA" = "cis-OPDA", "JA" = "JA", "OH_JA" = "OH-JA"))
#save plot
ggsave(
  filename = "cond_phyto_big.svg",
  plot = p2,
  path = "./Plots/02_phytohormones",
  scale = 1.5,
  width = 1200,
  height = 700,
  units = "px",
  dpi = 300)


# make a barplot for small-axis
p1 <- ggplot(data = subset(df_phyto, phytohormone %in% small_axis),
            mapping = aes(x = phytohormone, y = mean), colour= plant_pathway_induced)+
  geom_bar(aes(fill = plant_pathway_induced), stat = "identity",
           position = position_dodge(0.8), width = 0.8, color = "black")+
  geom_errorbar(aes(ymin = mean-se, ymax = mean+se, group = plant_pathway_induced),
                width = 0.2, position = position_dodge(0.8))+
  scale_fill_manual(values = color_pathway_induced)+
  geom_point(data = subset(conditioning_long, phytohormone %in% small_axis),
              mapping = (aes(phytohormone, concentration, fill = plant_pathway_induced)),
               shape = 21, size = 1.5, color = "black",
              position = position_jitterdodge(jitter.height = .1, jitter.width = .2))

p2 <- p1 + facet_wrap(~Experiment) +
            theme(strip.background = element_blank(),
                  strip.text.x = element_blank())+ #no label on top
              labs(y="(ng/g DW)", x=" ") +
              axis_looks +
              theme(panel.spacing = unit(0.2, "lines"))+
              theme(legend.position = "none")+
              scale_x_discrete(labels=c("sum_JA_Ile" = "JA-Ile", "OH_JA_Ile" = "OH-JA-Ile",
                                      "COOH_JA_Ile" = "COOH-JA-Ile"))

#save plot
ggsave(
  filename = "cond_phyto_small.svg",
  plot = p2,
  path = "./Plots/02_phytohormones",
  scale = 1.5,
  width = 1200,
  height = 700,
  units = "px",
  dpi = 300)

#clean
rm(p1,p2)
dev.off()

```

These two plots will be fused in one in illustrator

## Feedback

```{r}

#subset existing df to only feedback
phyto
long

#subsetting df to only conditioning
feedback <- phyto %>% 
                filter(Phase == "feedback")%>%
                subset(soil_inocula != "non-conditioned")%>%
                subset(ID != 140 & ID!= 146 & ID!= 159 & ID!= 163 & ID!= 208 & ID!= 215
                       & ID!= 220 & ID!= 222 & ID!= 224)%>% #removing those  with NA values
                droplevels()

#subsetting df to only conditioning
feedback_long <- long %>% 
                filter(Phase == "feedback")%>%
                subset(soil_inocula != "non-conditioned")%>%
                subset(ID != 140 & ID!= 146 & ID!= 159 & ID!= 163 & ID!= 208 & ID!= 215
                       & ID!= 220 & ID!= 222 & ID!= 224)%>%
                droplevels()

#change names of soil_inocula
levels(feedback$soil_inocula)
feedback <- feedback  %>%
            mutate(soil_inocula=factor(soil_inocula, labels=c("control_inocula", 
                                                              "JA_inocula",
                                                              "SA_inocula")))
levels(feedback_long$soil_inocula)
feedback_long <- feedback_long  %>%
                  mutate(soil_inocula=factor(soil_inocula, labels=c("control_inocula", 
                                                              "JA_inocula",
                                                              "SA_inocula")))
#change names of herbivory
levels(feedback$herbivory)
feedback <- feedback  %>%
            mutate(herbivory=factor(herbivory, levels=c("without", "caterpillar", "aphid")))

feedback <- feedback  %>%
            mutate(herbivory=factor(herbivory,
                                    labels=c("uninfested", "Mamestra", "Myzus")))

levels(feedback_long$herbivory)
feedback_long <- feedback_long  %>%
                  mutate(herbivory=factor(herbivory, levels=c("without", "caterpillar", "aphid")))

feedback_long <- feedback_long  %>%
                mutate(herbivory=factor(herbivory,
                                    labels=c("uninfested", "Mamestra", "Myzus")))

```

### PLSDA

```{r}

View(feedback)

str(feedback)
matrixdata<-feedback[,11:18]# keep only values
matrixdata <- as.matrix(matrixdata[,])# make it numeric
metadata<-feedback[,c("Experiment", "soil_inocula", "herbivory")]
#metadata <- metadata %>% mutate(joined=paste(Experiment, herbivory, sep="_"))
metadata <- factor(metadata$herbivory) #using only herbivory as explanatory variable

#PLS-DA 
plsda <-opls(matrixdata, metadata, scaleC = "pareto", permI = 100) #herbivory alone is a much better predictor 
#pareto removes the relative importance of large values
#but keeps the data structure intact.

#get summary
plot(plsda, typeVc = "x-score", parCexN = 1.5)
vip_scores <- getVipVn(plsda)

#let's remove those with VIP < 0.75
low_vip <- which(vip_scores < 0.75)
filtered_matrixdata <- matrixdata[,-low_vip]

#re-run PLSDA
plsda_filt <-opls(filtered_matrixdata, metadata, scaleC = "pareto", permI = 100)
plot(plsda_filt)

#get scores
scores<-plsda_filt@scoreMN
scores_df<-as.data.frame(scores) #make new df 

scores_df$p1 <- as.numeric(scores_df$p1)
scores_df$p2 <- as.numeric(scores_df$p2)

metadata<-feedback[,c("Experiment", "soil_inocula", "herbivory")]

scores_df2 <- scores_df %>% #add info
              mutate(Experiment = as.factor(metadata$Experiment))%>%
              mutate(soil_inocula = as.factor(metadata$soil_inocula))%>%
              mutate(herbivory=as.factor(metadata$herbivory))
```


Biplot
```{r}

#Plot the scores using ggplot2

str(scores_df2) #ok
ggpubr::show_point_shapes() #check

#plot
scoreplot<-ggplot(scores_df2, aes(x = p1, y = p2,
                                  fill=herbivory,#inside
                                  shape=soil_inocula, #shape
                                  color=Experiment))+ #outline
  geom_point(size = 4, stroke=0.5, alpha=0.8,
             aes(color=Experiment, fill=herbivory))+
  scale_shape_manual(values=c(21,22,24))+
  scale_color_manual(values = c("black", "white"))+ #outline
  scale_fill_manual(values=color_pathway_induced)+ #inside
  labs(x = "LV1 (60%)", y = "LV2 (21%)",
  color="Experiment", shape="herbivory", fill="Soil inocula")+
  axis_looks

#now add loadings
loadings<-plsda_filt@loadingMN
loadings_df<-as.data.frame(loadings)

loadings_df$p1 <- as.numeric(loadings_df$p1)
loadings_df$p2 <- as.numeric(loadings_df$p2)

#makes plot with arrows marking direction
hormone_names <- c("SA","JA","cis-OPDA","OH-JA")

#make biplot
Biplot <-scoreplot+
         geom_hline(yintercept=0, linetype='dotted', col = 'gray20',linewidth = 0.2 )+
         geom_vline(xintercept=0, linetype='dotted', col = 'gray20', linewidth = 0.2)+
         geom_segment(data = loadings_df, aes(x = 0, y = 0, xend = (p1*70), yend = (p2*70)),
                      arrow = arrow(length = unit(0.5, "picas"), type="closed"),
                      colour = "black", linewidth = 0.5, inherit.aes = FALSE)+
         geom_text(aes(x = (p1*70), y = (p2*70), label = hormone_names,
                angle = 0.45, hjust = -.1, vjust = .2),
                 data = loadings_df,
                 colour = "black", size=4, inherit.aes = FALSE)+
        #ylim(-60,60)+
        #xlim(-100,100)+
        axis_looks
         #theme(legend.position = "none")

#save plot
mapply(function(x)
ggsave(
  plot = last_plot(),
  filename =x,
  path = "./Plots/02_phytohormones",
  scale = 1.5,
  width = 1200,
  height = 700,
  units = "px",
  dpi = 300),
x=c("feedback_hormones_PLSDA.svg",
    "feedback_hormones_PLSDA.png"))

```



```{r}
#make same plot but colored by soil inocula
scoreplot<-ggplot(scores_df2, aes(x = p1, y = p2,
                                  fill=herbivory,#inside
                                  shape=soil_inocula, #shape
                                  color=Experiment))+ #outline
  geom_point(size = 4, stroke=0.5, alpha=0.8,
             aes(color=Experiment, fill=herbivory))+
  scale_shape_manual(values=c(21,22,24))+
  scale_color_manual(values = c("black", "white"))+ #outline
  scale_fill_manual(values=color_pathway_induced)+ #inside
  labs(x = "t1(36%)", y = "t2(34%)",
  color="Experiment", shape="Soil inocula", fill="Herbivory")+
  axis_looks

Biplot <-scoreplot+
         geom_hline(yintercept=0, linetype='dotted', col = 'gray20',linewidth = 0.2 )+
         geom_vline(xintercept=0, linetype='dotted', col = 'gray20', linewidth = 0.2)+
         geom_segment(data = loadings_df, aes(x = 0, y = 0, xend = (p1*70), yend = (p2*70)),
                      arrow = arrow(length = unit(0.5, "picas"), type="closed"),
                      colour = "black", linewidth = 0.5, inherit.aes = FALSE)+
         geom_text(aes(x = (p1*70), y = (p2*70), label = hormone_names,
                angle = 0.45, hjust = -.1, vjust = .2),
                 data = loadings_df,
                 colour = "black", size=4, inherit.aes = FALSE)+
        #ylim(-60,60)+
        #xlim(-100,100)+
        axis_looks
         #theme(legend.position = "none")

#save plot
ggsave(
  filename = "feedback_phyto_PLSDA_ByHerbivory_long.svg",
  plot = Biplot,
  path = "./Plots/02_phytohormones",
  scale = 1.5,
  width = 1000,
  height = 900,
  units = "px",
  dpi = 300)


```

Now, let's analyze the data with a PERMANOVA. First, data is auto-scaled to calculate a distance matrix from which a PERMANOVA is run to see whether there is an effect of plant pathway induced, the type of induction and the interaction of these two factors. 

### PERMANOVA

```{r}

#PERMANOVA
UVmatrixdata<- santaR:::scaling_UV(filtered_matrixdata) #for auto-scaling using st. deviation
DistanceMatrix<-dist(UVmatrixdata, method="euclidean")

#make it a df again
MetaData <-as.data.frame(metadata) 
MetaData[,c(1:3)]<- lapply(MetaData[,c(1:3)], as.factor)

#run PERMANOVA
permanova <- adonis2(DistanceMatrix~soil_inocula*herbivory*Experiment,
                      data=MetaData, permutations = 999, method="euclidean")

#make table for Word
#write names of factors
. <- c("Soil inocula (S)", "Herbivory (H)", "Experiment (E)", "S*H", "S*E", "H*E", "S*H*E",
       "Residual", "Total") #names of factors
permanova_table <- cbind (.,as.data.frame(permanova)) #make it a df

#keep 3 decimals after point for p values 
permanova_table[,c(3:6)] <- format(permanova_table[,c(3:6)], digits = 3, nsmall = 3)
permanova_table[,c(3:6)] <- lapply(permanova_table[,c(3:6)], as.numeric) #change it to numeric

#make it a nice_table
table <- nice_table(permanova_table,
              title = c("Table X. PERMANOVA for Plant Phytohormones", "Feedback phase",
                        "DistanceMatrix~soil_inocula*herbivory*Experiment"),
              note = "999 number of permutations")
#save it 
flextable::save_as_docx(table, path = "./Tables/02_phytohormones/permanova_hormones_feedback.docx")


```

### Individual Barplots

Boxplot with rawdata points, colored by plant_pathway_induced and facet wraped by Experiment + colored in the upper boxes. For this, first I need to make a summary table df. 

```{r}
#summary stats 


df_phyto <- feedback_long %>%
  group_by(Experiment, phytohormone, soil_inocula, herbivory) %>%
  summarise(
    n = n(),
    mean = mean(concentration, na.rm = TRUE),
    sd = sd(concentration, na.rm = TRUE)) %>%
    mutate(se = sd/sqrt(n))

df_phyto

```

Now, let's plot it. I'd like to have a plot with those hormones with high concentrations and then a second plot with a smaller axis for those of low concentrations so differences can be better appreciated.

```{r}

big_axis <- c("ABA", "SA", "cis_OPDA", "JA", "OH_JA")
small_axis <- c("sum_JA_Ile", "OH_JA_Ile","COOH_JA_Ile")


# make a barplot for big-axis
p1 <- ggplot(data = subset(feedback_long, phytohormone %in% big_axis),
            mapping = aes(x = soil_inocula, y = concentration), fill= soil_inocula)+
  geom_boxplot(aes(fill = soil_inocula))+
  #geom_bar(aes(fill = soil_inocula), stat = "identity",
   #        position = position_dodge(0.8), width = 0.8, color = "black")+
  #geom_errorbar(aes(ymin = mean-se, ymax = mean+se, group = soil_inocula),
  #              width = 0.2, position = position_dodge(0.8))+
  geom_point(data = subset(feedback_long, phytohormone %in% big_axis),
              mapping = (aes(soil_inocula, concentration, fill = soil_inocula)),
               shape = 21, size = 1.1, color = "black",
              position = position_jitterdodge(jitter.height = .1, jitter.width = .2))+
  scale_fill_manual(values = color_pathway_induced)

p2 <- p1 + facet_wrap(phytohormone+herbivory~Experiment, ncol=6) +
              labs(y="(ng/g DW)", x=" ") +
              axis_looks +
              theme(panel.spacing = unit(0.2, "lines"))+
              theme(legend.position = "none")+
              scale_x_discrete(labels=c("ABA" = "ABA", "SA" = "SA",
                                      "cis_OPDA" = "cis-OPDA", "JA" = "JA", "OH_JA" = "OH-JA"))
#save plot
ggsave(
  filename = "feed_phyto_big.svg",
  plot = p2,
  path = "./Plots/02_phytohormones",
  scale = 1.5,
  width = 1200,
  height = 1800,
  units = "px",
  dpi = 300)


# make a barplot for small-axis
p1 <- ggplot(data = subset(df_phyto, phytohormone %in% small_axis),
            mapping = aes(x = phytohormone, y = mean), fill= soil_inocula)+
  geom_bar(aes(fill = soil_inocula), stat = "identity",
           position = position_dodge(0.8), width = 0.8, color = "black")+
  geom_errorbar(aes(ymin = mean-se, ymax = mean+se, group = soil_inocula),
                width = 0.2, position = position_dodge(0.8))+
  geom_point(data = subset(feedback_long, phytohormone %in% small_axis),
              mapping = (aes(phytohormone, concentration, fill = soil_inocula)),
               shape = 21, size = 1.1, color = "black",
              position = position_jitterdodge(jitter.height = .1, jitter.width = .2))+
  scale_fill_manual(values = color_pathway_induced)

p2 <- p1 + facet_wrap(herbivory~Experiment, ncol=2) +
              labs(y="(ng/g DW)", x=" ") +
              axis_looks +
              theme(panel.spacing = unit(0.2, "lines"))+
              theme(legend.position = "none")+
              scale_x_discrete(labels=c("sum_JA_Ile" = "JA-Ile", "OH_JA_Ile" = "OH-JA-Ile",
                                      "COOH_JA_Ile" = "COOH-JA-Ile"))
#save plot
ggsave(
  filename = "feed_phyto_small.svg",
  plot = p2,
  path = "./Plots/02_phytohormones",
  scale = 1.5,
  width = 1200,
  height = 1800,
  units = "px",
  dpi = 300)

#clean
rm(p1,p2)
dev.off()

```

These two plots will be fused in one in illustrator




```{r}
save.image("./R_Environments/plant_hormones/plant_hormones.Rdata")
```

